---
title: "Functional Simulation"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(Matrix)
library(MASS)
library(fda)
library(lme4)
library(pedigreemm)
library(mvnfast)
library(ggplot2)
library(plotly)
```

## Functional Mixed-Effect Model

$$ \bf{Y} = \bf{Z^G \alpha} + \bf{Z^E \gamma} + \bf{\epsilon}$$ with the following distribution of the random vectors: \begin{align*}
\alpha &\sim N(\bf{0}, \bf{A \otimes C^G})\\

\gamma &\sim N(\bf{0}, \bf{I_N \otimes C^G})\\
\epsilon & \sim N(\bf{0}, \sigma^2_{res} *\bf{I_n})
\end{align*} assuming there are $N$ individuals and total $n$ measurements.

For the simulation study, we will fix a basis $\phi(t)$ for our functional space and true covariance matrices $\bf{C^G}$ and $\bf{C^E}$. Therefore the true genetic and environmental covariance functions which will be estimated are:
\begin{align*}
G(t,s) & = \phi(t)^T \ast \bf{C^G} \ast \phi(t)\\
E(t,s) & = \phi(t)^T \ast \bf{C^E} \ast \phi(t)
\end{align*}
Then we will generate a set of responses and fit the mixed-effect model to get the estimated covariance functions, which then will be compared with the true ones.

## Simulation process

Step 1: Fix functional basis, covariance matrices and residual variance. Here we will use 5 cubic B-spline basis.
```{r}
### use b-spline basis
basisObj <- create.bspline.basis(c(0,1), nbasis = 5, norder = 4)
### genetic covariance matrix
C_gen <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### environmental covariance matrix
C_env <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### residual variance
sigma2 <- 50
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width="150%"}
## Plot the true covariances functions
time_rang <- seq(0,1,length=10)
basis <- eval.basis(time_rang, basisObj) 
C_gen_fun <- basis %*% C_gen %*% t(basis)
C_env_fun <- basis %*% C_env %*% t(basis)
P_true <- C_gen_fun + C_env_fun

sim1 <- plot_ly(x = time_rang, y = time_rang, z = ~C_gen_fun, scene='scene1') 
sim1 <- sim1 %>% add_surface(showscale=FALSE)

sim2 <- plot_ly(x = time_rang, y = time_rang,z = ~C_env_fun, scene='scene2') 
sim2 <- sim2 %>% add_surface(showscale=FALSE)

sim3 <- plot_ly(x = time_rang, y = time_rang,z = ~P_true, scene='scene3') 
sim3 <- sim3 %>% add_surface(showscale=FALSE)

fig1 <- subplot(sim1, sim2, sim3) 
fig1 <- fig1 %>% layout(title = "Covariance Plot",
                              scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                           xaxis=list(title = "time"),
                                           yaxis =list(title = "time"),
                                           zaxis=list(range=c(-10,1600),
                                                      title = "Gen"),
                                           aspectmode='cube'),
                              scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                            xaxis=list(title = "time"),
                                            yaxis =list(title = "time"), 
                                            zaxis=list(range=c(-10,1600),
                                                       title = "Env"),
                                            aspectmode='cube'),
                              scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                            xaxis=list(title = "time"),
                                            yaxis =list(title = "time"), 
                                            zaxis=list(range=c(-10,1600),
                                                       title = "Phe"),
                                            aspectmode='cube'))

fig1
```

Step 2: Set the distribution of the random effect vectors $\bf{\alpha}$ and $\bf{\gamma}$.

Remark: we will assume in total there are 873 individuals so the same genetic relationship matrix $A$ computed in developing our mixed-effect model will be used in our simulation study. For now, assume each individual has 10 regular sampling points from the unit interval. We will use 3 (or 4) principal components to re-construct the covariance functions.

```{r include =FALSE}
### Calculate the genetic relationship matrix A (use the TC dataset)
setwd("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code")
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = 8730 # each individual have 10 observations, in total 8730

pos = df$id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]
```

```{r}
### reparameterised genetic covariance 
C_gen_para <- as(kronecker(A, C_gen), "dgCMatrix") 

### reparameterised environmental covariance 
I_N <- as(diag(873), "dgCMatrix")
C_env_para <- as(kronecker(I_N, C_env), "dgCMatrix")

### distribution of the genetic random vector
mu <- rep(0, dim(C_gen_para)[1])
alpha <- rmvn(n=50, mu = mu, sigma = C_gen_para) # here we generate 50 random vectors from multivariate normal distribution
gamma <- rmvn(n=50, mu = mu, sigma = C_env_para)

### distribution of the error vector
mu_res <- rep(0, 8730)
I_n <- as(diag(8730), "dgCMatrix")
res_cov <- as(sigma2 * I_n, "dgCMatrix") 
  
res <- rmvn(n=50, mu=mu_res, sigma = res_cov) # error vector
```

Step 3: Use the modular functions of `lme4()` to compute the random effect design matrices.  

```{r}
uniqueIds <- seq(1,873, length=873)
id <- rep(uniqueIds, each=10)
time_rang <- seq(0,1,length=10)
basis <- eval.basis(time_rang, basisObj)
b1 <- rep(basis[,1], times = 873)
b2 <- rep(basis[,2], times = 873)
b3 <- rep(basis[,3], times = 873)
b4 <- rep(basis[,4], times = 873)
b5 <- rep(basis[,5], times = 873)

y_pre <- rep(0, 8730) # dummy response values, will be updated 
df_pre <- data.frame(id = id, y_pre = y_pre, b1 = b1, b2 = b2, b3 = b3, b4 = b4, b5 = b5)

parsedFormula <- y_pre ~ (-1 + df_pre$b1 + df_pre$b2 + df_pre$b3 + df_pre$b4 + df_pre$b5 | df_pre$id) + 
  (-1 + df_pre$b1 + df_pre$b2 + df_pre$b3 + df_pre$b4 + df_pre$b5 | df_pre$id)

Z_pre <- t(lFormula(formula=parsedFormula, data=df_pre)$reTrms$Zt)

### update the genetic covariance matrix
L <- as(t(chol(A)), "dgCMatrix")
I5 <- as(diag(5), "dgCMatrix")
M <- kronecker(L, I5)
ZE <- Z_pre[,1:4365] 
ZG <- Z_pre[,1:4365] %*% M 
df_pre$y_pre <- as.vector(ZG %*% alpha[1,] + ZE %*% gamma[1,] + res[1,]) # random effects + residual
```

Step 4: Use `lm()` to calculate the fixed-effect. 

```{r}
f <- lm(y_pre ~ -1 + df_pre$b1 + df_pre$b2 + df_pre$b3 + df_pre$b4 + df_pre$b5, data = df_pre)
fix_ef <- f$coefficients[1] * df_pre$b1 + f$coefficients[2] * df_pre$b2 + f$coefficients[3] * df_pre$b3 +
  f$coefficients[4] * df_pre$b4 + f$coefficients[5] + df_pre$b5
```

Step 5: Generate response and reform into a dataframe for model-fitting.

```{r}
response <- as.vector(fix_ef + df_pre$y_pre) 
df_simu <- data.frame(id = id, response = response)
```

```{r echo=FALSE}
y_list <- split(response, df_simu$id)
plot(x=c(0,1), y=c(-300,300), type="n", xlab="time", ylab = "response")
for (i in 1:873){
  lines(time_rang, y_list[[i]], type = "l", col = i)
}
```

```{r include=FALSE}
fit_genetic_fmm <- function(formula, data, A, phi)
  {
  #'This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #'with a specified additive genetic relationship matrix A.
  #'In this particular format, we fit both fixed effects and random effects using the same
  #'basis functions (principal components obtained from running FPCA).
  #'
  #'@param fromula a two-sided linear formula object describing both the fixed-effects
  #'and random-effects of the model (as the same form used in lmer).
  #'@param data an data frame containing the variables named in formula.
  #'@param A a sparse matrix: an additive genetic relationship matrix 
  #'which models the genetic relationship in the dataset.
  #'@param phi functional basis: a matrix where each column represents a basis element.
  #'@return returns a fitted mixed-effect model
  
  # Random effect parameterisation
  require(lme4)
  require(Matrix)
  
  L <- as(t(chol(A)), "dgCMatrix") # cholesky decomposition of A
  p <- dim(phi)[2] # number of elements of the functional basis
  I_p <- as(diag(p), "dgCMatrix")
  M <- kronecker(L, I_p) # used to update the genetic design matrix Z_E = ZM
  
  # Fit mixed-effect model
  
  ## define the mixed-model formula
  fmmParsedForm <- lFormula(formula, data=data)
  
  ### Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[,1:dim(M)[1]] # environmental random-effect matrix
  ZG <- Z_pre[,1:dim(M)[1]] %*% M # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  ### Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun,fmmParsedForm) # update the objective function
  fmmOpitimize <- optimizeLmer(devfun=fmmDevFun)# update the optimisation module
  
  ### returns the mixed-effect model
  fmm <- mkMerMod(rho=environment(fmmDevFun),opt=fmmOpitimize, reTrms=fmmParsedForm$reTrms, fr=fmmParsedForm$fr)

  return(fmm)
}
```

## Fit simulated data
```{r}
## Data smoothing
timefine <- seq(0,1,length=25) # dense time grid
y_hat <- matrix(0, 25, 873)
lambda <- rep(0, 873)
for (i in 1:873){
  ss <- smooth.spline(time_rang, y_list[[i]], cv = FALSE)
  y_hat[,i] <- predict(ss, timefine)$y
  lambda[i] <- ss$lambda
}

matplot(timefine, y_hat, col = 1:873, type= "l") ### smoothed curves
```

```{r}
## FPCA
fpcaobj <- prcomp(x=t(y_hat), retx = TRUE, center = TRUE, rank. = 4)
pcs <- fpcaobj$rotation # eigen vectors as basis functions for moedel-fitting

summary(pcs)

pcs[,1] %*% pcs[,2]
```

```{r}
## Combine the smoothed data and basis functions to a dataframe
subjectID <- rep(unique(df_simu$id), each=25)
response_test <- c(y_hat)
basis1 <- rep(pcs[,1], times = 873)
basis2 <- rep(pcs[,2], times = 873)
basis3 <- rep(pcs[,3], times = 873)
basis4 <- rep(pcs[,4], times = 873)

df_test <- data.frame(id = subjectID, y= response_test, phi1 = basis1, phi2 = basis2, phi3 = basis3, phi4 = basis4) 

## Fit FMEM
fform <- y ~ -1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)

system.time(
 ft <-  fit_genetic_fmm(fform, df_test, A, pcs[,1:3])
) # user   system   elapsed

summary(ft)
```

```{r}
# Extract covariance function
vc <- VarCorr(ft)
CG <- vc[["df_test.id"]] ## genetic covariance
CE <- vc[["df_test.id.1"]] ## environmental covariance


### Convert to genetic covariance function
CG_fun <- pcs[,1:3] %*% CG %*% t(pcs[,1:3])
### environmental covariance function
CE_fun <- pcs[,1:3] %*% CE %*% t(pcs[,1:3])
### Phenotypic covariance function
P_fun <- CG_fun + CE_fun
```

```{r warning=FALSE, message=FALSE, echo=FALSE, out.width="150%"}
fig1 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun, scene='scene1') 
fig1 <- fig1 %>% add_surface(showscale=FALSE)

fig2 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun, scene='scene2') 
fig2 <- fig2 %>% add_surface(showscale=FALSE)

fig3 <- plot_ly(x = timefine, y = timefine,z = ~P_fun, scene='scene3') 
fig3 <- fig3 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR1 <- subplot(fig1, fig2, fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(title = "Phe"),
                                    aspectmode='cube'))

fig_RR1
```

### Covariance estimation when 4 PCs used as basis functions

```{r}
## Fit FMEM
fform1 <- y ~ -1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 | df_test$id) + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 | df_test$id)

system.time(
 ft1 <-  fit_genetic_fmm(fform1, df_test, A, pcs)
) # user   system   elapsed

summary(ft1)
```

```{r}
# Extract covariance function
vc1 <- VarCorr(ft1)
CG1 <- vc1[["df_test.id"]] ## genetic covariance
CE1 <- vc1[["df_test.id.1"]] ## environmental covariance


### Convert to genetic covariance function
CG_fun1 <- pcs %*% CG1 %*% t(pcs)
### environmental covariance function
CE_fun1 <- pcs %*% CE1 %*% t(pcs)
### Phenotypic covariance function
P_fun1 <- CG_fun1 + CE_fun1
```

```{r warning=FALSE, message=FALSE, echo=FALSE, out.width="150%"}
# individual plots
fig4 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun1, scene='scene1') 
fig4 <- fig4 %>% add_surface(showscale=FALSE)

fig5 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun1, scene='scene2') 
fig5 <- fig5 %>% add_surface(showscale=FALSE)

fig6 <- plot_ly(x = timefine, y = timefine,z = ~P_fun1, scene='scene3') 
fig6 <- fig6 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR2 <- subplot(fig4, fig5, fig6) 
fig_RR2 <- fig_RR2 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time"),
                                   zaxis=list(title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(title = "Phe"),
                                    aspectmode='cube'))

fig_RR2
```