---
title: "Model-fitting Efficiency"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
fit_genetic_fmm <- function(formula, data, A, phi)
  {
  #'This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #'with a specified additive genetic relationship matrix A.
  #'In this particular format, we fit both fixed effects and random effects using the same
  #'basis functions (principal components obtained from running FPCA).
  #'
  #'@param fromula a two-sided linear formula object describing both the fixed-effects
  #'and random-effects of the model (as the same form used in lmer).
  #'@param data an data frame containing the variables named in formula.
  #'@param A a sparse matrix: an additive genetic relationship matrix 
  #'which models the genetic relationship in the dataset.
  #'@param phi functional basis: a matrix where each column represents a basis element.
  #'@return returns a fitted mixed-effect model
  
  # Random effect parameterisation
  require(lme4)
  require(Matrix)
  
  L <- as(t(chol(A)), "dgCMatrix") # cholesky decomposition of A
  p <- dim(phi)[2] # number of elements of the functional basis
  I_p <- as(diag(p), "dgCMatrix")
  M <- kronecker(L, I_p) # used to update the genetic design matrix Z_E = ZM
  
  # Fit mixed-effect model
  
  ## define the mixed-model formula
  fmmParsedForm <- lFormula(formula, data=data)
  
  ### Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[,1:dim(M)[1]] # environmental random-effect matrix
  ZG <- Z_pre[,1:dim(M)[1]] %*% M # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  ### Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun,fmmParsedForm) # update the objective function
  fmmOpitimize <- optimizeLmer(devfun=fmmDevFun)# update the optimisation module
  
  ### returns the mixed-effect model
  fmm <- mkMerMod(rho=environment(fmmDevFun),opt=fmmOpitimize, reTrms=fmmParsedForm$reTrms, fr=fmmParsedForm$fr)

  return(fmm)
}
```

```{r include=FALSE}
convert_to_basisfunctions <- function(t, eigenvecs, tout) 
  {
  #'This function converts an eigenvector to an eigenfunction using interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  
  # Initialize an empty matrix to store eigenfunctions
  
  if (is.vector(eigenvecs) == TRUE){
    for (i in 1: length(eigenvecs)){
        eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
    }
  }
  
  else{
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  # Interpolate eigenvectors to the original time points
    
  return(eigen_functions)

}
```

## Import data
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(fdasrvf)
library(lme4)
library(pedigreemm)
library(ggplot2)
library(plotly)

setwd("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code")
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = length(df$id) # n = 6860 observations
age_list <- split(df$x,df$id)
trait_list <- split(df$trait,df$id)

## Rescale time interval to [0,1]
## x = (x -min(x))/(max(x) - min(x))
age_list_new <- list()
for (i in 1:N){
  age_list_new[[i]] = (age_list[[i]]-min(age_list[[i]]))/(max(age_list[[i]])-min(age_list[[i]]))
}

df$x_rescaled <- unsplit(age_list_new,df$id)
```

## Data smoothing

Here we smooth growth curves on the logarithmic scale and then take exponential to recover body mass. This imposes positive smoothing. Smoothing parameter $\lambda_i$ for each curve is selected by GCV and we restrict $\lambda \le 10^{-4}$.

```{r}
agefine <- seq(0,1,length=100) # dense time grid
logmass <- matrix(0,100,N) # store smooth log growth curves
pred_mass <- matrix(0,100,N) # store the smoothed mass recovered by taking exp
lam <- rep(0,N) # smoothing parameter used for each log growth curve

for (i in 1:N){
  ss_logmass <- smooth.spline(age_list_new[[i]], log(trait_list[[i]]), cv=FALSE,
                              all.knots=TRUE)
  logmass[,i] <- predict(ss_logmass, agefine)$y
  pred_mass[,i] <- exp(predict(ss_logmass,agefine)$y)
  lam[i] <- ss_logmass$lambda
}

large_lam <- which(lam > 1e-4)
mass_adjusted <- matrix(0,100, length(large_lam))
for ( i in 1: length(large_lam)){
  ss_new <- smooth.spline(age_list_new[[large_lam[i]]], log(trait_list[[large_lam[i]]]), 
                          all.knots = TRUE, lambda = 1e-4) # restrict lambda <=1e-4
  mass_adjusted[,i] <- exp(predict(ss_new, agefine)$y)
}
colnames(mass_adjusted) <- as.character(large_lam)

## Let reform the smoothed growth curves into a matrix
trait_hat <- pred_mass 
for (i in 1: length(large_lam)){
  trait_hat[,large_lam[i]] <- mass_adjusted[,i]
}
```

```{r echo=FALSE}
## Plot the smoothed growth curves
par(mfrow=c(1,2))

plot(c(0,1), c(0,400), xlab="Time", ylab="Mass", main="Growth Plot (rescaled time)", type="n")
for (i in 1:N){
  lines(age_list_new[[i]], trait_list[[i]],type="l", col=i)
}

matplot(agefine, trait_hat, col=1:N, type = "l", xlab="Time", ylab="Mass", main="Smoothed Growth Plot")
```

## Curve Alignment

Align the growth curves using the r package $\textbf{fdasrvf}$ and extract the mean from the aligned curves.

```{r warning=FALSE, message=FALSE}
aligned_mass_process <- time_warping(f=trait_hat, time=agefine)
aligned_mass_curve <- aligned_mass_process$fn
aligned_mean <- aligned_mass_process$fmean
warping_funs <- aligned_mass_process$warping_functions
```

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(c(0,1), c(0,1), type = 'n',xlab = 'Time', 
     ylab = 'warping functions')
for (i in 1:N){
  lines(agefine, warping_funs[,i], type="l" , col=i)
}


plot(c(0,1), c(0,400), type = 'n',xlab = 'Time', 
     ylab = 'Mass', main = 'Aligned Growth Plot')
for (i in 1:N){
  lines(agefine, aligned_mass_curve[,i], type="l",col=i)
}
lines(agefine, aligned_mean, lwd = 3.0, col="red")
legend("bottomright", legend="mean", lwd=3.0, col="red")
```

## Functional Principal Component Analysis

We run FPCA on the aligned data and the first three principal components will be used as the basis functions for our model.

```{r}
fpcaobj_mass <- prcomp(x=t(aligned_mass_curve), retx = TRUE, center = TRUE, rank. = 3)
eigen_mass <- fpcaobj_mass$rotation # eigen vectors
```

```{r echo=FALSE}
par(mfrow=c(3,1))
for (i in 1:3) {
  plot(agefine, eigen_mass[, i], type = "l", 
       xlab = "time", ylab = "",
       main = paste("Growth Curve Principal Component", i))
}

## Test for orthogonality
eigen_mass[,1] %*% eigen_mass[,2] 
eigen_mass[,1] %*% eigen_mass[,3] 
eigen_mass[,2] %*% eigen_mass[,3] 
```

## Fit Functional Mixed-Effect Model

We calculate the additive genetic relationship matrix $\bf{A}$ and fit both fixed and random effects using the same set of functional basis.

```{r warning=FALSE, message=FALSE}
pos = df$id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]
```

Here we compare three different ways to fit data to our genetic model:

1.  Fit raw data:

We need to interpolate the discrete eigen vectors to eigen functions and evaluate eigen functions at the original time points. This will be done using the function `convert_to_basisfunctions`.

```{r }
### define the basis functions (eigenfunctions)
phi_list <- list() 
# create an empty list which stores eigenfunctions for 873 subjects
# evaluated at the original time points.

for (i in 1:N){
  phi <- convert_to_basisfunctions(t = agefine, eigenvecs = eigen_mass[,1:3],
                                   tout = age_list_new[[i]])
  phi_list[[i]] <- phi
}

phi <- do.call(rbind,phi_list)
colnames(phi) <- c("phi1", "phi2", "phi3")

### update the dataframe to include basis functions
df1 <- cbind(df,phi)

### use a fixed regression of the same form of random regression 
fmmForm1 <- trait ~ df1$phi1 + df1$phi2 + df1$phi3 + (-1 + df1$phi1 + df1$phi2 + df1$phi3 | df1$id) + 
  (-1 + df1$phi1 + df1$phi2 + df1$phi3 | df1$id) 

system.time(
  ff1 <- fit_genetic_fmm(fmmForm1, df1, A, eigen_mass)
) # user   system  elapsed

summary(ff1)
```

Extract the covariance matrices and convert them to covariance functions.

```{r}
VC1 <- VarCorr(ff1)
CG1 <- VC1[["df1.id"]]
CE1 <- VC1[["df1.id.1"]]

### Convert to genetic covariance function
CG_fun1 <- eigen_mass%*% CG1 %*% t(eigen_mass)
### environmental covariance function
CE_fun1 <- eigen_mass %*% CE1 %*% t(eigen_mass)
### Phenotypic covariance function
P_fun1 <- CG_fun1 + CE_fun1
```

```{r warning=FALSE, message=FALSE, echo=FALSE, out.width = "150%"}
# individual plots
fig1 <- plot_ly(x = agefine, y = agefine, z = ~CG_fun1, scene='scene1') 
fig1 <- fig1 %>% add_surface(showscale=FALSE)

fig2 <- plot_ly(x = agefine, y = agefine,z = ~CE_fun1, scene='scene2') 
fig2 <- fig2 %>% add_surface(showscale=FALSE)

fig3 <- plot_ly(x = agefine, y = agefine,z = ~P_fun1, scene='scene3') 
fig3 <- fig3 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR1 <- subplot(fig1, fig2, fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range=c(-10,1200),
                                              title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                   title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                            title = "Phe"),
                                    aspectmode='cube'))

fig_RR1
```

2.  Fit the smoothed data on the original time points.

We interpolate the smoothed data at the original measurement points.

```{r}
trait_hat_list <- list()

for (i in 1:N){
  inter_trait <- convert_to_basisfunctions(t = agefine, eigenvecs = trait_hat[,i],
                                   tout = age_list_new[[i]])
  trait_hat_list[[i]] <- inter_trait
}

df2 <- data.frame(id = df$id, trait = unsplit(trait_hat_list, df$id), phi)

fmmForm2 <- trait ~ df2$phi1 + df2$phi2 + df2$phi3 + (-1 + df2$phi1 + df2$phi2 + df2$phi3 | df2$id) + 
  (-1 + df2$phi1 + df2$phi2 + df2$phi3 | df2$id) 

system.time(
  ff2 <- fit_genetic_fmm(fmmForm2, df2, A, eigen_mass)
) # user   system  elapsed

summary(ff2)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width = "150%"}
VC2 <- VarCorr(ff2)
CG2 <- VC2[["df2.id"]]
CE2 <- VC2[["df2.id.1"]]

### Convert to genetic covariance function
CG_fun2 <- eigen_mass %*% CG2 %*% t(eigen_mass)
### environmental covariance function
CE_fun2 <- eigen_mass %*% CE2 %*% t(eigen_mass)
### Phenotypic covariance function
P_fun2 <- CG_fun2 + CE_fun2

# individual plots
fig4 <- plot_ly(x = agefine, y = agefine, z = ~CG_fun2, scene='scene1') 
fig4 <- fig4 %>% add_surface(showscale=FALSE)

fig5 <- plot_ly(x = agefine, y = agefine,z = ~CE_fun2, scene='scene2') 
fig5 <- fig5 %>% add_surface(showscale=FALSE)

fig6 <- plot_ly(x = agefine, y = agefine,z = ~P_fun2, scene='scene3') 
fig6 <- fig6 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR2 <- subplot(fig4, fig5, fig6) 
fig_RR2 <- fig_RR2 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time"),
                                   zaxis=list(range=c(-10,1200),
                                           title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(range=c(-10,1200),
                                            title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(range=c(-10,1200),
                                            title = "Phe"),
                                    aspectmode='cube'))

fig_RR2
```

3. Fit the smoothed data on the dense time grid.

```{r}
### Reform the aligned data to a dataframe for model fitting
subjectID <- rep(unique(df$id), each=100)
trait_pred <- c(aligned_mass_curve)
basis1 <- rep(eigen_mass[,1], times = 873)
basis2 <- rep(eigen_mass[,2], times = 873)
basis3 <- rep(eigen_mass[,3], times = 873)
new_df <- data.frame(subjectID, trait_pred, basis1, basis2, basis3)
names(new_df) <- c("subjectID", "trait_hat", "basis1", "basis2", "basis3")

fmeFormula <- trait_hat ~ new_df$basis1 + new_df$basis2 + new_df$basis3 + 
  (-1 + new_df$basis1 + new_df$basis2 + new_df$basis3 | new_df$subjectID) + 
  (-1 + new_df$basis1 + new_df$basis2 + new_df$basis3 | new_df$subjectID)

system.time(
  fit_sameBasis <- fit_genetic_fmm(formula= fmeFormula, data=new_df, A = A, phi = eigen_mass)
) 
 # user   system   elapsed

summary(fit_sameBasis)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width = "150%"}
VC3 <- VarCorr(fit_sameBasis)
CG3 <- VC3[["new_df.subjectID"]]
CE3 <- VC3[["new_df.subjectID.1"]]

### Convert to genetic covariance function
CG_fun3 <- eigen_mass %*% CG3 %*% t(eigen_mass)
### environmental covariance function
CE_fun3 <- eigen_mass %*% CE3 %*% t(eigen_mass)
### Phenotypic covariance function
P_fun3 <- CG_fun3 + CE_fun3

# individual plots
fig7 <- plot_ly(x = agefine, y = agefine, z = ~CG_fun3, scene='scene1') 
fig7 <- fig7 %>% add_surface(showscale=FALSE)

fig8 <- plot_ly(x = agefine, y = agefine,z = ~CE_fun3, scene='scene2') 
fig8 <- fig8 %>% add_surface(showscale=FALSE)

fig9 <- plot_ly(x = agefine, y = agefine,z = ~P_fun3, scene='scene3') 
fig9 <- fig9 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR3 <- subplot(fig7, fig8, fig9) 
fig_RR3 <- fig_RR3 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range=c(-10,1200),
                                              title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                             title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                             title = "Phe"),
                                    aspectmode='cube'))

fig_RR3
```
