---
title: "Simulation"
output: html_document
date: "2024-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
fit_genetic_fmm <- function(formula, data, A, phi)
  {
  #'This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #'with a specified additive genetic relationship matrix A.
  #'In this particular format, we fit both fixed effects and random effects using the same
  #'basis functions (principal components obtained from running FPCA).
  #'
  #'@param fromula a two-sided linear formula object describing both the fixed-effects
  #'and random-effects of the model (as the same form used in lmer).
  #'@param data an data frame containing the variables named in formula.
  #'@param A a sparse matrix: an additive genetic relationship matrix 
  #'which models the genetic relationship in the dataset.
  #'@param phi functional basis: a matrix where each column represents a basis element.
  #'@return returns a fitted mixed-effect model
  
  # Random effect parameterisation
  require(lme4)
  require(Matrix)
  
  L <- as(t(chol(A)), "dgCMatrix") # cholesky decomposition of A
  p <- dim(phi)[2] # number of elements of the functional basis
  I_p <- as(diag(p), "dgCMatrix")
  M <- kronecker(L, I_p) # used to update the genetic design matrix Z_E = ZM
  
  # Fit mixed-effect model
  
  ## define the mixed-model formula
  fmmParsedForm <- lFormula(formula, data=data)
  
  ### Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[,1:dim(M)[1]] # environmental random-effect matrix
  ZG <- Z_pre[,1:dim(M)[1]] %*% M # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  ### Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun,fmmParsedForm) # update the objective function
  fmmOpitimize <- optimizeLmer(devfun=fmmDevFun)# update the optimisation module
  
  ### returns the mixed-effect model
  fmm <- mkMerMod(rho=environment(fmmDevFun),opt=fmmOpitimize, reTrms=fmmParsedForm$reTrms, fr=fmmParsedForm$fr)

  return(fmm)
}
```

```{r include=FALSE}
convert_to_basisfunctions <- function(t, eigenvecs, tout) 
  {
  #'This function converts an eigenvector to an eigenfunction using interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  
  # Initialize an empty matrix to store eigenfunctions
  
  if (is.vector(eigenvecs) == TRUE){
    for (i in 1: length(eigenvecs)){
        eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
    }
  }
  
  else{
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  # Interpolate eigenvectors to the original time points
    
  return(eigen_functions)

}
```

```{r include=FALSE}
library(Matrix)
library(MASS)
library(fda)
library(lme4)
library(pedigreemm)
library(mvnfast)
library(ggplot2)
library(plotly)
library(fBasics)
library(latex2exp)
```

## Functional Mixed-Effect Model

\begin{align*}
\bf{Y} &= \bf{X \beta} + \bf{Z^G \alpha} + \bf{Z^E \gamma} + \bf{\epsilon}\\
&= \bf{X \beta} + \bf{Z u} + \bf{\epsilon}
\end{align*} where $\bf{Z} = [\bf{Z^G}, \bf{Z^E}]$, $\bf{u} = [\bf{\alpha}, \bf{\beta}]^T$ with the following distribution of the random vectors: \begin{align*}
\alpha &\sim N(\bf{0}, \bf{A \otimes C^G})\\
\gamma &\sim N(\bf{0}, \bf{I_N \otimes C^E})\\
\epsilon & \sim N(\bf{0}, \sigma^2_{res} *\bf{I_n})
\end{align*} assuming there are $N$ individuals and total $n$ measurements.

For the simulation study, we will fix a basis $\phi(t)$ for our functional space and true covariance matrices $\bf{C^G}$ and $\bf{C^E}$. Therefore the true genetic and environmental covariance functions which will be estimated are: \begin{align*}
G(t,s) & = \phi(t)^T \ast \bf{C^G} \ast \phi(t)\\
E(t,s) & = \phi(t)^T \ast \bf{C^E} \ast \phi(t)
\end{align*} Then we will generate a set of responses and fit the mixed-effect model to get the estimated covariance functions $\hat{G}(t,s)$ and $\hat{E}(t,s)$, which then will be compared with the true ones.

## Simulation process

Step 1: Use 5 cubic B-splines as model basis. Fix equal covariance for both genetic and environmental effects.

Remark: Later we will use principal components as basis to fit the mixed-effect model. Here we assume the number of basis to construct the covariance functions is larger than the number of principal components to re-construct the covariance.

```{r}
set.seed(123)

### use b-spline basis
basisObj <- create.bspline.basis(c(0,1), nbasis = 5, norder = 4)
### genetic covariance matrix
C_gen <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### environmental covariance matrix
C_env <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### residual variance
sigma2 <- 50
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width="120%"}
## Plot the true covariances functions
time_rang <- seq(0,1,length=10)
basis <- eval.basis(time_rang, basisObj) 
C_gen_fun <- basis %*% C_gen %*% t(basis)
C_env_fun <- basis %*% C_env %*% t(basis)
P_true <- C_gen_fun + C_env_fun

trace_ratio_true <- tr(C_gen_fun) / tr(C_env_fun)

sim1 <- plot_ly(x = time_rang, y = time_rang, z = ~C_gen_fun, scene='scene1') 
sim1 <- sim1 %>% add_surface(showscale=FALSE)

sim2 <- plot_ly(x = time_rang, y = time_rang,z = ~C_env_fun, scene='scene2') 
sim2 <- sim2 %>% add_surface(showscale=FALSE)

fig1 <- subplot(sim1, sim2) 
fig1 <- fig1 %>% layout(title = "Covariance Plot",
                              scene = list(domain=list(x=c(0,0.45),y=c(0.25,1)),
                                           xaxis=list(title = "time"),
                                           yaxis =list(title = "time"),
                                           zaxis=list(range=c(-10,800),
                                                      title = "Gen"),
                                           aspectmode='cube'),
                              scene2 = list(domain=list(x=c(0.50,0.95),y=c(0.25,1)),
                                            xaxis=list(title = "time"),
                                            yaxis =list(title = "time"), 
                                            zaxis=list(range=c(-10,800),
                                                       title = "Env"),
                                            aspectmode='cube'))

fig1
```

Step 2: Set the distribution of the random effect vectors $\bf{\alpha}$ and $\bf{\gamma}$.

Remark: we will assume there are 873 individuals so the same genetic relationship matrix $A$ computed from the actual data set will be used in our simulation study. For now, assume each individual has 10 regular sampling points from the unit interval. We will use 3 (and 4) principal components to re-construct the covariance functions (so we see how the number of principal components affects model result).

```{r include =FALSE}
### Calculate the genetic relationship matrix A (use the TC dataset)
setwd("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code")
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = 8730 # each individual have 10 observations, in total 8730

pos = df$id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]
```

```{r}
### reparameterised genetic covariance 
C_gen_para <- as(kronecker(A, C_gen), "dgCMatrix") 

### reparameterised environmental covariance 
I_N <- as(diag(873), "dgCMatrix")
C_env_para <- as(kronecker(I_N, C_env), "dgCMatrix")


### distribution of the random vectors 
mu <- rep(0, dim(C_gen_para)[1])
alpha <- rmvn(n=50, mu, C_gen_para) # generated 50 repsonses
gamma <- rmvn(n=50, mu, C_env_para)

### distribution of the error vector
mu_res <- rep(0, 8730)
I_n <- as(diag(8730), "dgCMatrix")
res_cov <- as(sigma2 * I_n, "dgCMatrix") 
  
res <- rmvn(n=50, mu=mu_res, sigma = res_cov) # error vector
```

Step 3: Generate curve data.

Each subject's curve data is influenced by a combination of random effects across the basis functions plus the measurement errors. Here we manually map random effects back to the individual level, considering the Kronecker product structure used in generating $\bf{\alpha}$ and $\bf{\beta}$.

```{r}
rand_eff <- matrix(0, n, 50) # matrix where each column contains the random effects of one simulated sample
errors <- matrix(0, n, 50) # matrix where each column contains the residual of one simulated sample
for (j in 1:50){
  rand_vec <- numeric(0)
  for (i in seq(1,length(alpha[j,]), by = 5)){
    rand <- basis %*% alpha[j,i:(i+4)] + basis %*% gamma[j,i:(i+4)]
    rand_vec <- c(rand_vec, rand)
  }
  rand_eff[,j] <- rand_vec 
}

for (j in 1:50){
  err_vec <- numeric(0)
  for (k in seq(1, length(res[j,]), by = 10)){
    err <- res[j,k:(k+9)]
    err_vec <- c(err_vec, err)
  }
  errors[,j] <- err_vec
}

Y <- rand_eff + errors # matrix where each column contains the simulated data of one sample
```

```{r echo=FALSE}
uniqueIds <- seq(1,873, length=873)
id <- rep(uniqueIds, each=10)
y_list <- split(Y[,1], id)
plot(x=c(0,1), y=c(-150,150), type="n", xlab="time", ylab = "response")
for (i in 1:873){
  lines(time_rang, y_list[[i]], type = "l", col = i)
}
```

## Fit simulated data

```{r}
## Data smoothing
timefine <- seq(0,1,length=100) # dense time grid
y_hat <- matrix(0, 100, 873)

for (i in 1:873){
  ss <- smooth.spline(time_rang, y_list[[i]], cv = FALSE)
  y_hat[,i] <- predict(ss, timefine)$y
}

matplot(timefine, y_hat, col = 1:873, type= "l") ### smoothed curves
```

```{r}
## FPCA
### fit smoothed data at original time measurement points
trait <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)

fpcaobj <- prcomp(x=t(trait), retx = TRUE, center = TRUE, rank. = 4)
pcs <- fpcaobj$rotation # eigen vectors as basis functions for moedel-fitting

pcs[,1] %*% pcs[,2]
```

```{r}
## Combine the smoothed data and basis functions to a dataframe
response_test <- c(trait)
basis1 <- rep(pcs[,1], times = 873)
basis2 <- rep(pcs[,2], times = 873)
basis3 <- rep(pcs[,3], times = 873)
basis4 <- rep(pcs[,4], times = 873)

df_test <- data.frame(id = id, y= response_test, phi1 = basis1, phi2 = basis2, phi3 = basis3, phi4 = basis4) 

## Fit FMEM
fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
  (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)

system.time(
 ft <-  fit_genetic_fmm(fform, df_test, A, pcs[,1:3])
) # user   system   elapsed

summary(ft)
```

```{r}
# Extract covariance function
vc <- VarCorr(ft)
CG <- vc[["df_test.id"]] ## genetic covariance
CE <- vc[["df_test.id.1"]] ## environmental covariance


### Convert to genetic covariance function
CG_fun <- pcs[,1:3] %*% CG %*% t(pcs[,1:3])
### environmental covariance function
CE_fun <- pcs[,1:3] %*% CE %*% t(pcs[,1:3])
### Phenotypic covariance function
P_fun <- CG_fun + CE_fun

### Error
gen_err1 <- norm((C_gen_fun - CG_fun), type = "F")
paste("Genetic error:",gen_err1)
env_err1 <- norm((C_env_fun - CE_fun), type = "F")
paste("Environment error:",env_err1)
```

```{r warning=FALSE, message=FALSE, echo=FALSE,out.width="120%"}
fig1 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun, scene='scene1') 
fig1 <- fig1 %>% add_surface(showscale=FALSE)

fig2 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun, scene='scene2') 
fig2 <- fig2 %>% add_surface(showscale=FALSE)

fig_RR1 <- subplot(fig1, fig2) 
fig_RR1 <- fig_RR1 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.45),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range=c(-10,900),title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.50,0.95),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,900),title = "Env"),
                                    aspectmode='cube'))

fig_RR1
```

So far we only test our model on one simulated response. Let us repeat the process on the rest simulated data and measure the total errors (measured by Frobenius norm) on genetic and environmental covariance function for each simulation.

```{r eval=FALSE}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id <- rep(uniqueIds, each=10)
y_hat <- matrix(0, 100, 873)
gen_err3 <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3 <- rep(0,50) ## store the trace of genetic covariance function
tr_env3 <- rep(0,50) ## store the trace of environmental covariance function
```

```{r eval=FALSE}
system.time(
for (j in 41:50){
  ### data smoothing
  Y_list <- split(Y[,j], id)
  for (i in 1:873){
    ss <- smooth.spline(time_rang, Y_list[[i]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 873)
  basis2 <- rep(pcs[,2], times = 873)
  basis3 <- rep(pcs[,3], times = 873)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3[j] <- tr(CG_fun)
  tr_env3[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r eval=FALSE}
save(tr_gen3,file = "tr_gen3.Rdata")
save(tr_env3,file = "tr_env3.Rdata")
save(gen_err3,file = "gen_err3.Rdata")
save(env_err3,file = "env_err3.Rdata")
```

```{r include=FALSE}
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/tr_env3.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/tr_gen3.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/env_err3.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/gen_err3.Rdata")
```

```{r echo=FALSE}
par(mfrow = c(1,1))
errors_list <- list(gen_err3, env_err3)

# Generate the boxplot
boxplot(errors_list,
        xlab = "",
        ylab = "Error",
        col = c("blue", "green"),
        names = c("Genetic", "Environmental"),
        main = "Errors when using 3 PCs")
```

```{r echo=FALSE}
plot(c(1,50), c(0.5,2.5),type="n", xlab="", ylab = TeX(r'(${tr(gen)/tr(env)}$)'), main = "Trace ratio when using 3 PCs")
x_axis <- seq(1,50,by=1)
for (j in 1:50){
  lines(x_axis[j],(tr_gen3[j]/tr_env3[j]), type = "p")
}
abline(h=1,col="blue")

paste("average trace ratio:",sum((tr_gen3/tr_env3))/50)
```

### Simulation using 4 PCs

```{r eval = FALSE}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id <- rep(uniqueIds, each=10)
y_hat <- matrix(0, 100, 873)
gen_err4 <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err4 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm

tr_gen4 <- rep(0,50) ## store the trace of genetic covariance function
tr_env4 <- rep(0,50) ## store the trace of environmental covariance function

```

```{r eval = FALSE}
system.time(
for (j in 1:5){
  ### data smoothing
  Y_list <- split(Y[,j], id)
  for (i in 1:873){
    ss <- smooth.spline(time_rang, Y_list[[i]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 4)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 873)
  basis2 <- rep(pcs[,2], times = 873)
  basis3 <- rep(pcs[,3], times = 873)
  basis4 <- rep(pcs[,4], times = 873)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3, phi4 = basis4) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 +
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + df_test$phi4 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen4[j] <- tr(CG_fun)
  tr_env4[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err4[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err4[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r eval=FALSE}
save(tr_gen4,file = "tr_gen4.Rdata")
save(tr_env4,file = "tr_env4.Rdata")
save(gen_err4,file = "gen_err4.Rdata")
save(env_err4,file = "env_err4.Rdata")
```

```{r include=FALSE}
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/tr_env4.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/tr_gen4.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/env_err4.Rdata")
load("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code/Simulation_results_50groups/gen_err4.Rdata")
```

```{r echo=FALSE}
par(mfrow = c(1,1))
errors_list <- list(gen_err4, env_err4)

# Generate the boxplot
boxplot(errors_list,
        xlab = "",
        ylab = "Error",
        col = c("blue", "green"),
        names = c("Genetic", "Environmental"),
        main = "Errors when using 4 PCs")
```

```{r echo=FALSE}
plot(c(1,50), c(0.5,2.5),type="n", xlab="", ylab = TeX(r'(${tr(gen)/tr(env)}$)'), main = "Trace ratio when using 4 PCs")
x_axis <- seq(1,50,by=1)
for (j in 1:50){
  lines(x_axis[j],(tr_gen4[j]/tr_env4[j]), type = "p")
}
abline(h=1,col="blue")

paste("average trace ratio:",sum((tr_gen4/tr_env4))/50)
```