---
title: "Subsampling"
output: html_document
date: "2024-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
fit_genetic_fmm <- function(formula, data, A, phi)
  {
  #'This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #'with a specified additive genetic relationship matrix A.
  #'In this particular format, we fit both fixed effects and random effects using the same
  #'basis functions (principal components obtained from running FPCA).
  #'
  #'@param fromula a two-sided linear formula object describing both the fixed-effects
  #'and random-effects of the model (as the same form used in lmer).
  #'@param data an data frame containing the variables named in formula.
  #'@param A a sparse matrix: an additive genetic relationship matrix 
  #'which models the genetic relationship in the dataset.
  #'@param phi functional basis: a matrix where each column represents a basis element.
  #'@return returns a fitted mixed-effect model
  
  # Random effect parameterisation
  require(lme4)
  require(Matrix)
  
  L <- as(t(chol(A)), "dgCMatrix") # cholesky decomposition of A
  p <- dim(phi)[2] # number of elements of the functional basis
  I_p <- as(diag(p), "dgCMatrix")
  M <- kronecker(L, I_p) # used to update the genetic design matrix Z_E = ZM
  
  # Fit mixed-effect model
  
  ## define the mixed-model formula
  fmmParsedForm <- lFormula(formula, data=data)
  
  ### Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[,1:dim(M)[1]] # environmental random-effect matrix
  ZG <- Z_pre[,1:dim(M)[1]] %*% M # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  ### Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun,fmmParsedForm) # update the objective function
  fmmOpitimize <- optimizeLmer(devfun=fmmDevFun)# update the optimisation module
  
  ### returns the mixed-effect model
  fmm <- mkMerMod(rho=environment(fmmDevFun),opt=fmmOpitimize, reTrms=fmmParsedForm$reTrms, fr=fmmParsedForm$fr)

  return(fmm)
}
```

```{r include=FALSE}
convert_to_basisfunctions <- function(t, eigenvecs, tout) 
  {
  #'This function converts an eigenvector to an eigenfunction using interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  
  # Initialize an empty matrix to store eigenfunctions
  
  if (is.vector(eigenvecs) == TRUE){
    for (i in 1: length(eigenvecs)){
        eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
    }
  }
  
  else{
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  # Interpolate eigenvectors to the original time points
    
  return(eigen_functions)

}
```

```{r include=FALSE}
library(Matrix)
library(MASS)
library(fda)
library(lme4)
library(pedigreemm)
library(mvnfast)
library(ggplot2)
library(plotly)
library(fBasics)
```

### Simulated data

```{r include =FALSE}
### Calculate the genetic relationship matrix A (use the TC dataset)
setwd("D:/KCL_2023-2027_PhD/Year 1/Genetics/FMEMs-quantitative-genetics/R_code")
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = 10*N # each individual have 10 observations, in total 8730

pos = df$id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]
```

```{r}
set.seed(123)

### use b-spline basis
basisObj <- create.bspline.basis(c(0,1), nbasis = 5, norder = 4)
### genetic covariance matrix
C_gen <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### environmental covariance matrix
C_env <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### residual variance
sigma2 <- 50
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width="120%}
## Plot the true covariances functions
time_rang <- seq(0,1,length=10)
basis <- eval.basis(time_rang, basisObj) 
C_gen_fun <- basis %*% C_gen %*% t(basis)
C_env_fun <- basis %*% C_env %*% t(basis)
P_true <- C_gen_fun + C_env_fun

sim1 <- plot_ly(x = time_rang, y = time_rang, z = ~C_gen_fun, scene='scene1') 
sim1 <- sim1 %>% add_surface(showscale=FALSE)

sim2 <- plot_ly(x = time_rang, y = time_rang,z = ~C_env_fun, scene='scene2') 
sim2 <- sim2 %>% add_surface(showscale=FALSE)

fig1 <- subplot(sim1, sim2) 
fig1 <- fig1 %>% layout(title = "Covariance Plot",
                              scene = list(domain=list(x=c(0,0.45),y=c(0.25,1)),
                                           xaxis=list(title = "time"),
                                           yaxis =list(title = "time"),
                                           zaxis=list(range=c(-10,800),
                                                      title = "Gen"),
                                           aspectmode='cube'),
                              scene2 = list(domain=list(x=c(0.50,0.95),y=c(0.25,1)),
                                            xaxis=list(title = "time"),
                                            yaxis =list(title = "time"), 
                                            zaxis=list(range=c(-10,800),
                                                       title = "Env"),
                                            aspectmode='cube'))

fig1
```

```{r}
### reparameterised genetic covariance 
C_gen_para <- as(kronecker(A, C_gen), "dgCMatrix") 

### reparameterised environmental covariance 
I_N <- as(diag(873), "dgCMatrix")
C_env_para <- as(kronecker(I_N, C_env), "dgCMatrix")

mu <- rep(0, dim(C_gen_para)[1])
alpha <- rmvn(n=50, mu, C_gen_para)
gamma <- rmvn(n=50, mu, C_env_para)

### distribution of the error vector
mu_res <- rep(0, 8730)
I_n <- as(diag(8730), "dgCMatrix")
res_cov <- as(sigma2 * I_n, "dgCMatrix") 
  
res <- rmvn(n=50, mu=mu_res, sigma = res_cov) # error vector
```

```{r}
rand_eff <- matrix(0, n, 50)
errors <- matrix(0, n, 50)
for (j in 1:50){
  rand_vec <- numeric(0)
  for (i in seq(1, length(alpha[j,]), by = 5)){
    rand <- basis %*% alpha[j, i:(i+4)] + basis %*% gamma[j, i:(i+4)]
    rand_vec <- c(rand_vec,rand)
  }
  rand_eff[,j] <- rand_vec
}

for (j in 1:50){
  err_vec <- numeric(0)
  for (k in seq(1, length(res[j,]), by = 10)){
    err <- res[j,k:(k+9)]
    err_vec <- c(err_vec,err)
  }
  errors[,j] <- err_vec
}

Y <- rand_eff + errors
```

In the previous simulation, we generate data for all individuals. Here we do sub-sampling of size 100, 400 and 600 according to the additive genetic relationship matrix $\bf{A}$. Recall that $\bf{A}$ is a positive definite matrix with each term $A_{ij}$ representing how closely related between individuals $i$ and $j$. For example, $A_{ij} = 0.5$ means the two individuals have both sire and dam in common and $A_{ij} = 0.25$ means the two individuals either have a sire or a dam in common.

Case 1: Select individuals as diverse as possible

Step1: Choose an individual at random;

Step2: Choose subsequent individuals with the lowest average relation score.

We first sample 100 individuals.

```{r}
diverse_sampling <- function(A, sample_size){
  #'This function maximises genetic diversity in a sub-sample.
  #'
  #'@param A: additive genetic relationship matrix
  #'@param sample_size: number of individuals in a sub-sample.
  #'
  #'@return return offprings IDs
  
  N <- dim(A)[1] # number of total individuals
  indices <- 1:N 
  selected <- sample(indices, 1)  # Start by selecting one individual randomly
  remaining <- setdiff(indices, selected)  # Remaining individuals to choose from
  
  while (length(selected) < sample_size) {
    candidate_selected <- NULL
    optimal_relation_score <- Inf
    
    # Iterate through remaining candidates
    for (candidate in remaining) {
      candidate_relation_score <- mean(A[candidate, selected, drop = FALSE])
      if (candidate_relation_score < optimal_relation_score) {
        candidate_selected <- candidate
        optimal_relation_score <- candidate_relation_score
      }
    }
    
  # Update the selected and remaining lists
    selected <- c(selected, candidate_selected)
    remaining <- setdiff(remaining, candidate_selected)
  }
  
  return(selected)
}
```

```{r}
set.seed(123)
sample100 <- sort(diverse_sampling(A,100))
sample100 
```
We need to calculate the additive genetic relationship matrix for this 100-subsample.

```{r}
pos100 <- pos[sample100] # extract ids for all subjects
sire100 <- df$sire[FirstUniqueIdPos[sample100]] # extract ids for sire
dam100 <- df$dam[FirstUniqueIdPos[sample100]] # extract ids for dam

pede <- editPed(sire = sire100, dam = dam100, label = pos100)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A100 <- getA(ped)[130:229,130:229]
```

Fit the mixed-effect model on the selected sub-sample:

```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample100, each=10)
y_hat <- matrix(0, 100, 100)
gen_err3_100 <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_100 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_100 <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_100 <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:100){
    ss <- smooth.spline(time_rang, Y_list[[sample100[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 100)
  basis2 <- rep(pcs[,2], times = 100)
  basis3 <- rep(pcs[,3], times = 100)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A100, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_100[j] <- tr(CG_fun)
  tr_env3_100[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_100[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_100[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_100,file = "tr_gen3_100.Rdata")
save(tr_env3_100,file = "tr_env3_100.Rdata")
save(gen_err3_100,file = "gen_err3_100.Rdata")
save(env_err3_100,file = "env_err3_100.Rdata")
```

Let us sample 400 individuals.

```{r}
sample400 <- sort(diverse_sampling(A,400))
```

```{r}
pos400 <- pos[sample400] # extract ids for all subjects
sire400 <- df$sire[FirstUniqueIdPos[sample400]] # extract ids for sire
dam400 <- df$dam[FirstUniqueIdPos[sample400]] # extract ids for dam

pede <- editPed(sire = sire400, dam = dam400, label = pos400)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A400 <- getA(ped)[163:562,163:562]
```

```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample400, each=10)
y_hat <- matrix(0, 100, 400)
gen_err3_400 <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_400 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_400 <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_400 <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:400){
    ss <- smooth.spline(time_rang, Y_list[[sample400[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 400)
  basis2 <- rep(pcs[,2], times = 400)
  basis3 <- rep(pcs[,3], times = 400)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A400, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_400[j] <- tr(CG_fun)
  tr_env3_400[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_400[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_400[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_400,file = "tr_gen3_400.Rdata")
save(tr_env3_400,file = "tr_env3_400.Rdata")
save(gen_err3_400,file = "gen_err3_400.Rdata")
save(env_err3_400,file = "env_err3_400.Rdata")
```

```{r}
sample600 <- sort(diverse_sampling(A,600))
```

```{r}
pos600 <- pos[sample600] # extract ids for all subjects
sire600 <- df$sire[FirstUniqueIdPos[sample600]] # extract ids for sire
dam600 <- df$dam[FirstUniqueIdPos[sample600]] # extract ids for dam

pede <- editPed(sire = sire600, dam = dam600, label = pos600)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A600 <- getA(ped)[163:762,163:762]
```

```{r include=FALSE}
save(sample100,file = "sample100.Rdata")
save(sample400,file = "sample400.Rdata")
save(sample600,file = "sample600.Rdata")
save(A100,file = "A100.Rdata")
save(A400,file = "A400.Rdata")
save(A600,file = "A600.Rdata")
```


```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample600, each=10)
y_hat <- matrix(0, 100, 600)
gen_err3_600 <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_600 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_600 <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_600 <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:600){
    ss <- smooth.spline(time_rang, Y_list[[sample600[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 600)
  basis2 <- rep(pcs[,2], times = 600)
  basis3 <- rep(pcs[,3], times = 600)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A600, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_600[j] <- tr(CG_fun)
  tr_env3_600[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_600[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_600[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_600,file = "tr_gen3_600.Rdata")
save(tr_env3_600,file = "tr_env3_600.Rdata")
save(gen_err3_600,file = "gen_err3_600.Rdata")
save(env_err3_600,file = "env_err3_600.Rdata")
```

Case2: Select individuals as related as possible.

Step1: Choose an individual at random;

Step2: Choose subsequent individuals with the highest average relation score.

```{r}
related_sampling <- function(A, sample_size){
  #'This function minimises genetic diversity in a sub-sample.
  #'
  #'@param A: additive genetic relationship matrix
  #'@param sample_size: number of individuals in a sub-sample.
  #'
  #'@return return selected individuals' IDs
  
  N <- dim(A)[1] # number of total individuals
  indices <- 1:N 
  selected <- sample(indices, 1)  # Start by selecting one individual randomly
  remaining <- setdiff(indices, selected)  # Remaining individuals to choose from
  
  while (length(selected) < sample_size) {
    candidate_selected <- NULL
    optimal_relation_score <- -Inf
    
    # Iterate through remaining candidates
    for (candidate in remaining) {
      candidate_relation_score <- mean(A[candidate, selected, drop = FALSE])
      if (candidate_relation_score > optimal_relation_score) {
        candidate_selected <- candidate
        optimal_relation_score <- candidate_relation_score
      }
    }
    
    # Update the selected and remaining lists
    selected <- c(selected, candidate_selected)
    remaining <- setdiff(remaining, candidate_selected)
  }
  
  return(selected)
}
```

```{r}
set.seed(123)
sample100R <- sort(related_sampling(A,100))
sample400R <- sort(related_sampling(A,400))
sample600R <- sort(related_sampling(A,600))
``` 

```{r}
sample100R
```

```{r include=FALSE}
save(sample100R,file = "sample100R.Rdata")
save(sample400R,file = "sample400R.Rdata")
save(sample600R,file = "sample600R.Rdata")
```
 
```{r}
## calculating the relation matrix for 100-subsample
pos100R <- pos[sample100R] # extract ids for all subjects
sire100R <- df$sire[FirstUniqueIdPos[sample100R]] # extract ids for sire
dam100R <- df$dam[FirstUniqueIdPos[sample100R]] # extract ids for dam

pede <- editPed(sire = sire100R, dam = dam100R, label = pos100R)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A100R <- getA(ped)[24:123,24:123]
```

```{r}
A100_R <- as.matrix(A100R)
```

```{r}
## calculating the relation matrix for 400-subsample
pos400R <- pos[sample400R] # extract ids for all subjects
sire400R <- df$sire[FirstUniqueIdPos[sample400R]] # extract ids for sire
dam400R <- df$dam[FirstUniqueIdPos[sample400R]] # extract ids for dam

pede <- editPed(sire = sire400R, dam = dam400R, label = pos400R)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A400R <- getA(ped)[76:475,76:475]
```

```{r}
## calculating the relation matrix for 600-subsample
pos600R <- pos[sample600R] # extract ids for all subjects
sire600R <- df$sire[FirstUniqueIdPos[sample600R]] # extract ids for sire
dam600R <- df$dam[FirstUniqueIdPos[sample600R]] # extract ids for dam

pede <- editPed(sire = sire600R, dam = dam600R, label = pos600R)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A600R <- getA(ped)[110:709,110:709]
```

```{r include=FALSE}
save(A100R,file = "A100R.Rdata")
save(A400R,file = "A400R.Rdata")
save(A600R,file = "A600R.Rdata")
```

Fit the mixed-effect model on the selected sub-sample:

We run model on 100 sub-samples.

```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample100R, each=10)
y_hat <- matrix(0, 100, 100)
gen_err3_100R <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_100R <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_100R <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_100R <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:100){
    ss <- smooth.spline(time_rang, Y_list[[sample100R[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 100)
  basis2 <- rep(pcs[,2], times = 100)
  basis3 <- rep(pcs[,3], times = 100)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A100R, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_100R[j] <- tr(CG_fun)
  tr_env3_100R[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_100R[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_100R[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_100R,file = "tr_gen3_100R.Rdata")
save(tr_env3_100R,file = "tr_env3_100R.Rdata")
save(gen_err3_100R,file = "gen_err3_100R.Rdata")
save(env_err3_100R,file = "env_err3_100R.Rdata")
```

Let us repeat the process on 400 individuals.

```{r}
pos400 <- pos[sample400] # extract ids for all subjects
sire400 <- df$sire[FirstUniqueIdPos[sample400]] # extract ids for sire
dam400 <- df$dam[FirstUniqueIdPos[sample400]] # extract ids for dam

pede <- editPed(sire = sire400, dam = dam400, label = pos400)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A400 <- getA(ped)[163:562,163:562]
```

```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample400R, each=10)
y_hat <- matrix(0, 100, 400)
gen_err3_400R <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_400R <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_400R <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_400R <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:400){
    ss <- smooth.spline(time_rang, Y_list[[sample400R[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 400)
  basis2 <- rep(pcs[,2], times = 400)
  basis3 <- rep(pcs[,3], times = 400)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A400R, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_400R[j] <- tr(CG_fun)
  tr_env3_400R[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_400R[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_400R[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_400R,file = "tr_gen3_400R.Rdata")
save(tr_env3_400R,file = "tr_env3_400R.Rdata")
save(gen_err3_400R,file = "gen_err3_400R.Rdata")
save(env_err3_400R,file = "env_err3_400R.Rdata")
```

Now we fit model with 600 individuals

```{r}
pos600 <- pos[sample600] # extract ids for all subjects
sire600 <- df$sire[FirstUniqueIdPos[sample600]] # extract ids for sire
dam600 <- df$dam[FirstUniqueIdPos[sample600]] # extract ids for dam

pede <- editPed(sire = sire600, dam = dam600, label = pos600)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A600 <- getA(ped)[163:762,163:762]
```

```{r include=FALSE}
save(sample100,file = "sample100.Rdata")
save(sample400,file = "sample400.Rdata")
save(sample600,file = "sample600.Rdata")
save(A100,file = "A100.Rdata")
save(A400,file = "A400.Rdata")
save(A600,file = "A600.Rdata")
```


```{r}
time_rang <- seq(0,1,length=10)
timefine <- seq(0,1,length=100) # dense time grid
uniqueIds <- seq(1,873, length=873)
id_873 <- rep(uniqueIds, each=10)
id <- rep(sample600R, each=10)
y_hat <- matrix(0, 100, 600)
gen_err3_600R <- rep(0, 50) ## store the error on genetic covariance in Frobenius norm
env_err3_600 <- rep(0, 50) ## store the error on environmental covariance in Frobenius norm
tr_gen3_600R <- rep(0,50) ## store the trace of genetic covariance function
tr_env3_600R <- rep(0,50) ## store the trace of environmental covariance function
```

```{r}
system.time(
for (j in 1:50){
  ### data smoothing
  Y_list <- split(Y[,j], id_873)
  for (i in 1:600){
    ss <- smooth.spline(time_rang, Y_list[[sample600R[i]]], cv = FALSE)
    y_hat[,i] <- predict(ss, timefine)$y
  }
  
  ### fit smoothed data at original time measurement points
  y <- convert_to_basisfunctions(t = timefine, eigenvecs = y_hat,
                                 tout = time_rang)
  
  ### FPCA
  fpcaobj <- prcomp(x=t(y), retx = TRUE, center = TRUE, rank. = 3)
  pcs <- fpcaobj$rotation
  
  ## Combine the smoothed data and basis functions to a dataframe
  response_test <- c(y)
  basis1 <- rep(pcs[,1], times = 600)
  basis2 <- rep(pcs[,2], times = 600)
  basis3 <- rep(pcs[,3], times = 600)
  
  df_test <- data.frame(id = id, y= response_test, phi1 = basis1, 
                        phi2 = basis2, phi3 = basis3) 
  
  ### Fit FMEM
  fform <- y ~ 1 + df_test$phi1 + df_test$phi2 + df_test$phi3 + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id) + 
    (-1 + df_test$phi1 + df_test$phi2 + df_test$phi3 | df_test$id)
  
  ft <-  fit_genetic_fmm(fform, df_test, A600R, pcs)
  
  ### Estimated covariance
  vc <- VarCorr(ft)
  CG <- vc[["df_test.id"]] # genetic covariance
  CE <- vc[["df_test.id.1"]] # environmental covariance
  
  CG_fun <- pcs %*% CG %*% t(pcs) # estimated gen cov function
  CE_fun <- pcs %*% CE %*% t(pcs) # estimated env cov function
  
  tr_gen3_600R[j] <- tr(CG_fun)
  tr_env3_600R[j] <- tr(CE_fun)
  
  ### Error on genetic and environmental covariance
  gen_err3_600R[j] <- norm((C_gen_fun - CG_fun), type = "F")
  env_err3_600R[j] <- norm((C_env_fun - CE_fun), type = "F")
}
)
```

```{r include=FALSE}
save(tr_gen3_600R,file = "tr_gen3_600R.Rdata")
save(tr_env3_600R,file = "tr_env3_600R.Rdata")
save(gen_err3_600R,file = "gen_err3_600R.Rdata")
save(env_err3_600R,file = "env_err3_600R.Rdata")
```
 