---
title: "Subsampling"
output: html_document
date: "2024-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
fit_genetic_fmm <- function(formula, data, A, phi)
  {
  #'This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #'with a specified additive genetic relationship matrix A.
  #'In this particular format, we fit both fixed effects and random effects using the same
  #'basis functions (principal components obtained from running FPCA).
  #'
  #'@param fromula a two-sided linear formula object describing both the fixed-effects
  #'and random-effects of the model (as the same form used in lmer).
  #'@param data an data frame containing the variables named in formula.
  #'@param A a sparse matrix: an additive genetic relationship matrix 
  #'which models the genetic relationship in the dataset.
  #'@param phi functional basis: a matrix where each column represents a basis element.
  #'@return returns a fitted mixed-effect model
  
  # Random effect parameterisation
  require(lme4)
  require(Matrix)
  
  L <- as(t(chol(A)), "dgCMatrix") # cholesky decomposition of A
  p <- dim(phi)[2] # number of elements of the functional basis
  I_p <- as(diag(p), "dgCMatrix")
  M <- kronecker(L, I_p) # used to update the genetic design matrix Z_E = ZM
  
  # Fit mixed-effect model
  
  ## define the mixed-model formula
  fmmParsedForm <- lFormula(formula, data=data)
  
  ### Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[,1:dim(M)[1]] # environmental random-effect matrix
  ZG <- Z_pre[,1:dim(M)[1]] %*% M # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  ### Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun,fmmParsedForm) # update the objective function
  fmmOpitimize <- optimizeLmer(devfun=fmmDevFun)# update the optimisation module
  
  ### returns the mixed-effect model
  fmm <- mkMerMod(rho=environment(fmmDevFun),opt=fmmOpitimize, reTrms=fmmParsedForm$reTrms, fr=fmmParsedForm$fr)

  return(fmm)
}
```

```{r include=FALSE}
convert_to_basisfunctions <- function(t, eigenvecs, tout) 
  {
  #'This function converts an eigenvector to an eigenfunction using interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  
  # Initialize an empty matrix to store eigenfunctions
  
  if (is.vector(eigenvecs) == TRUE){
    for (i in 1: length(eigenvecs)){
        eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
    }
  }
  
  else{
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  # Interpolate eigenvectors to the original time points
    
  return(eigen_functions)

}
```

```{r include=FALSE}
library(Matrix)
library(MASS)
library(fda)
library(lme4)
library(pedigreemm)
library(mvnfast)
library(ggplot2)
library(plotly)
library(fBasics)
```

### Simulated data

```{r include =FALSE}
### Calculate the genetic relationship matrix A (use the TC dataset)
setwd("H:/FMEMs-quantitative-genetics/R_code")
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = 8730 # each individual have 10 observations, in total 8730

pos = df$id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]
```

```{r}
set.seed(123)

### use b-spline basis
basisObj <- create.bspline.basis(c(0,1), nbasis = 5, norder = 4)
### genetic covariance matrix
C_gen <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### environmental covariance matrix
C_env <- matrix(c(750, 10 ,130, 80, 250,
              10, 800, 30, 15, 40,
              130, 30, 700, 50, 130,
              80, 15, 50, 420, 50,
              250, 40, 130, 50, 330), nrow = 5, byrow = T)
### residual variance
sigma2 <- 50
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
## Plot the true covariances functions
time_rang <- seq(0,1,length=10)
basis <- eval.basis(time_rang, basisObj) 
C_gen_fun <- basis %*% C_gen %*% t(basis)
C_env_fun <- basis %*% C_env %*% t(basis)
P_true <- C_gen_fun + C_env_fun

sim1 <- plot_ly(x = time_rang, y = time_rang, z = ~C_gen_fun, scene='scene1') 
sim1 <- sim1 %>% add_surface(showscale=FALSE)

sim2 <- plot_ly(x = time_rang, y = time_rang,z = ~C_env_fun, scene='scene2') 
sim2 <- sim2 %>% add_surface(showscale=FALSE)

fig1 <- subplot(sim1, sim2) 
fig1 <- fig1 %>% layout(title = "Covariance Plot",
                              scene = list(domain=list(x=c(0,0.45),y=c(0.25,1)),
                                           xaxis=list(title = "time"),
                                           yaxis =list(title = "time"),
                                           zaxis=list(range=c(-10,800),
                                                      title = "Gen"),
                                           aspectmode='cube'),
                              scene2 = list(domain=list(x=c(0.50,0.95),y=c(0.25,1)),
                                            xaxis=list(title = "time"),
                                            yaxis =list(title = "time"), 
                                            zaxis=list(range=c(-10,800),
                                                       title = "Env"),
                                            aspectmode='cube'))

fig1
```

```{r}
### reparameterised genetic covariance 
C_gen_para <- as(kronecker(A, C_gen), "dgCMatrix") 

### reparameterised environmental covariance 
I_N <- as(diag(873), "dgCMatrix")
C_env_para <- as(kronecker(I_N, C_env), "dgCMatrix")

mu <- rep(0, dim(C_gen_para)[1])
alpha <- rmvn(n=50, mu, C_gen_para)
gamma <- rmvn(n=50, mu, C_env_para)

### distribution of the error vector
mu_res <- rep(0, 8730)
I_n <- as(diag(8730), "dgCMatrix")
res_cov <- as(sigma2 * I_n, "dgCMatrix") 
  
res <- rmvn(n=50, mu=mu_res, sigma = res_cov) # error vector
```

```{r}
rand_eff <- matrix(0, n, 50)
errors <- matrix(0, n, 50)
for (j in 1:50){
  rand_vec <- numeric(0)
  for (i in seq(1, length(alpha[j,]), by = 5)){
    rand <- basis %*% alpha[j, i:(i+4)] + basis %*% gamma[j, i:(i+4)]
    rand_vec <- c(rand_vec,rand)
  }
  rand_eff[,j] <- rand_vec
}

for (j in 1:50){
  err_vec <- numeric(0)
  for (k in seq(1, length(res[j,]), by = 10)){
    err <- res[j,k:(k+9)]
    err_vec <- c(err_vec,err)
  }
  errors[,j] <- err_vec
}

Y <- rand_eff + errors
```

In the previous simulation, we generate data for all individuals. Here we do sub-sampling according to their genetic relation. For example, we can select individuals with full/half siblings and individuals not related at all. Then run our mixed-effect model and examine how reliable the model is in terms of estimating covariance under various genetic relations.

Case1: we sample 100 individuals with 70% full/half siblings and 30% not related.

```{r}
Id_matrix <- matrix(0, 873, 3)#
Id_matrix[,1] <- seq(1,873,by=1)
Id_matrix[,2] <- sire_id
Id_matrix[,3] <- dam_id
```


```{r}
check_genetic_relation(Id_matrix[1,], Id_matrix[2,])
check_genetic_relation(Id_matrix[1,], Id_matrix[78,])
```


```{r}
# Initialize lists to store selected individuals
related_individuals <- list()
unrelated_individuals <- list()

# Iterate through each pair of individuals and check relatedness
for (i in 1:872) {
  for (j in (i + 1):873) {
    if (check_genetic_relation(Id_matrix[i,], Id_matrix[j,]) == TRUE) {
      related_individuals[[length(related_individuals) + 1]] <- c(Id_matrix[i,1], Id_matrix[j,1])
    } else {
      unrelated_individuals[[length(unrelated_individuals) + 1]] <- c(Id_matrix[i,1], Id_matrix[j,1])
    }
  }
}
```

```{r}
# Sample 70% related individuals and 30% unrelated individuals
num_related <- 70
num_unrelated <- 30
```