---
title: "Model-fitting Efficiency"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
convert_to_basisfunctions <- function(t, eigenvecs, tout) 
{
  #'This function converts an eigenvector to an eigenfunction using linear interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  # Check if eigenvecs is a vector or matrix
  if (is.vector(eigenvecs)) {
    eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
  } else {
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  
  return(eigen_functions)
}

fit_genetic_fmm <- function(formula, data, A, nbasis, control = lmerControl()) {
  #' This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #' with a specified additive genetic relationship matrix A.
  #' In this particular format, we fit both fixed effects and random effects using the same
  #' basis functions (principal components obtained from running FPCA).
  #'
  #' @param formula a two-sided linear formula object describing both the fixed-effects
  #' and random-effects of the model (as the same form used in lmer).
  #' @param data a data frame containing the variables named in formula.
  #' @param A a sparse matrix: an additive genetic relationship matrix 
  #' which models the genetic relationship in the dataset.
  #' @param nbasis numeber: number of basis used to fit the random effects
  #' @param control lmerControl object: control parameters for the optimizer.
  #' @return returns a fitted mixed-effect model
  
  # Load required packages
  library(lme4)
  library(Matrix)
  
  # Cholesky decomposition of A
  LA <- as(t(chol(A)), "sparseMatrix")
  I_p <- as(diag(nbasis), "sparseMatrix")
  MA <- kronecker(LA, I_p) # used to update the genetic design matrix Z_G = ZM
  
  # Define the mixed-model formula
  fmmParsedForm <- lFormula(formula=formula, data = data, control = control)
  
  # Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[, 1:dim(MA)[1]] # environmental random-effect matrix
  ZG <- Z_pre[, 1:dim(MA)[1]] %*% MA # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  # Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun, fmmParsedForm) # update the objective function
  
  fmmOptimize <- optimizeLmer(devfun = fmmDevFun, control = control) # update the optimisation module
  
  # Return the mixed-effect model
  fmm <- mkMerMod(rho = environment(fmmDevFun), opt = fmmOptimize, reTrms = fmmParsedForm$reTrms, fr = fmmParsedForm$fr)
  
  return(fmm)
}

```

## Import data
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(fdasrvf)
library(lme4)
library(pedigreemm)
library(ggplot2)
library(plotly)

TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

## calculate genetic relationship matrix
id <- df$id
FirstUniqueIdPos <- which(duplicated(id) == FALSE)
pos = id[FirstUniqueIdPos] # extract ids for all subjects
sire_id = df$sire[FirstUniqueIdPos] # extract ids for sire
dam_id = df$dam[FirstUniqueIdPos] # extract ids for dam

pede <- editPed(sire = sire_id, dam = dam_id, label = pos)
ped<- with(pede, pedigree(label=label, sire=sire, dam=dam))
A <- getA(ped)[163:1035,163:1035]

## Rescale time interval to [0,1]
N = length(FirstUniqueIdPos) # N = 873 subjects
n = length(id) # n = 6860 observations
age_list <- split(df$x,id)
trait_list <- split(df$trait,id)

### x = (x -min(x))/(max(x) - min(x))
age_list_new <- list()
for (i in 1:N){
  age_list_new[[i]] = (age_list[[i]]-min(age_list[[i]]))/(max(age_list[[i]])-min(age_list[[i]]))
}

df$x_rescaled <- unsplit(age_list_new,id)
df$logtrait <- log10(df$trait)
id <- df$id

```

## Data smoothing

Here we smooth growth curves on the logarithmic scale and then take exponential to recover body mass. This imposes positive smoothing. Smoothing parameter $\lambda_i$ for each curve is selected by GCV and we restrict $\lambda \le 10^{-4}$.

```{r}
timefine <- seq(0,1,length=100)
mass_smoothed <-list()
pred_mass_fine <- matrix(0,100,N) # store the smoothed mass predicted on the dense grid
pred_logmass_fine <- matrix(0,100,N) # store the smoothed logmass
lam <- rep(0,N) # smoothing parameter used for each log growth curve

for (i in 1:N) {
  ss_logmass <- smooth.spline(age_list_new[[i]], log10(trait_list[[i]]), cv=FALSE, all.knots=TRUE)
  
  # Check if lambda is greater than 1e-4
  if (ss_logmass$lambda > 1e-4) {
    # Redo smoothing with lambda set to 1e-4
    ss_logmass <- smooth.spline(age_list_new[[i]], log10(trait_list[[i]]), lambda=1e-4, all.knots=TRUE)
  }
  
  mass_smoothed[[i]] <- 10^(ss_logmass$y)
  pred_logmass_fine[,i] <- predict(ss_logmass, timefine)$y
  pred_mass_fine[,i] <- 10^(predict(ss_logmass, timefine)$y) ## predict smoothed data on a regular dense grid
  lam[i] <- ss_logmass$lambda
}

```

```{r echo=FALSE}
## Plot the smoothed growth curves
par(mfrow=c(1,2))

plot(c(0,1), c(0,400), xlab="Time", ylab="Mass", main="Growth Plot", type="n")
for (i in 1:N){
  lines(age_list_new[[i]], trait_list[[i]],type="l", col=i)
}

matplot(timefine, pred_mass_fine, col=1:N, type = "l", xlab="Time", ylab="Mass", main="Smoothed Growth Plot")
```

## Curve Alignment

Align the growth curves using the r package $\textbf{fdasrvf}$ and extract the mean from the aligned curves.

```{r warning=FALSE, message=FALSE}
aligned_mass_process <- time_warping(pred_mass_fine, timefine)
aligned_mass_curve <- aligned_mass_process$fn
aligned_mass_mean <- aligned_mass_process$fmean
warping_mass_funs <- aligned_mass_process$warping_functions
```

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(c(0,1), c(0,1), type = 'n',xlab = 'Time', 
     ylab = 'warping functions')
for (i in 1:N){
  lines(timefine, warping_mass_funs[,i], type="l" , col=i)
}


plot(c(0,1), c(0,400), type = 'n',xlab = 'Time', 
     ylab = 'Mass', main = 'Aligned Growth Plot')
for (i in 1:N){
  lines(timefine, aligned_mass_curve[,i], type="l",col=i)
}
lines(timefine, aligned_mass_mean, lwd = 3.0, col="red")
legend("bottomright", legend="mean", lwd=3.0, col="red")
```

## Functional Principal Component Analysis

We run FPCA on the aligned data and the first three principal components will be used as the basis functions for our model.

```{r}
fpcaobj_mass <- prcomp(x=t(aligned_mass_curve), retx = TRUE, center = TRUE, rank. = 2)
pcs_mass <- fpcaobj_mass$rotation # eigen vectors
```

```{r echo=FALSE}
par(mfrow=c(2,1))
for (i in 1:2) {
  plot(timefine, pcs_mass[, i], type = "l", 
       xlab = "time", ylab = "",
       main = paste("Growth Curve Principal Component", i))
}

## Test for orthogonality
pcs_mass[,1] %*% pcs_mass[,2] 
```

## Fit Functional Mixed-Effect Model

We fit both fixed and random effects using the same set of functional basis.

Here we compare three different ways to fit data to our genetic model:

1.  Fit raw aligned data:

We need to interpolate the warping functions back to original sampling points and register raw curves. 

```{r }
aligned_trait <- list()
gamma_mass <- list()
for (i in 1:N){
  gamma_mass_inter <- convert_to_basisfunctions(timefine, warping_mass_funs[,i], age_list_new[[i]])
  gamma_mass[[i]] <- gamma_mass_inter
  aligned_trait[[i]] <- warp_f_gamma(trait_list[[i]], age_list_new[[i]], gamma_mass_inter)
}

### Prepare model basis
phi_mass_list <- list() 
# create an empty list which stores eigenfunctions for 873 subjects
# evaluated at the original time points.

for (i in 1:N){
  phi_mass <- convert_to_basisfunctions(t = timefine, eigenvecs = pcs_mass,
                                           tout = age_list_new[[i]])
  phi_mass_list[[i]] <- phi_mass
}

phi_mass <- do.call(rbind,phi_mass_list)
colnames(phi_mass) <- c("phi1", "phi2")

df_mass1 <- data.frame(id = id, trait = unsplit(aligned_trait,id), phi_mass)

fmmForm1 <- trait ~ -1 + df_mass1$phi1 + df_mass1$phi2 + 
  (-1 + df_mass1$phi1 + df_mass1$phi2 | df_mass1$id) + 
  (-1 + df_mass1$phi1 + df_mass1$phi2 | df_mass1$id) 

system.time(
  ff1 <- fit_genetic_fmm(fmmForm1, df_mass1, A, 2)
) # user   system  elapsed

summary(ff1)

isSingular(ff1) # TRUE singular fit, tol = 1e-4
```

Extract the covariance matrices and convert them to covariance functions.

```{r}
VC1 <- VarCorr(ff1)
CG1 <- VC1[["df_mass1.id"]]
CE1 <- VC1[["df_mass1.id.1"]]

### Convert to genetic covariance function
CG_fun1 <- pcs_mass%*% CG1 %*% t(pcs_mass)
### environmental covariance function
CE_fun1 <- pcs_mass %*% CE1 %*% t(pcs_mass)
### Phenotypic covariance function
P_fun1 <- CG_fun1 + CE_fun1
```

```{r warning=FALSE, message=FALSE, echo=FALSE, out.width = "150%"}
# individual plots
fig1 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun1, scene='scene1') 
fig1 <- fig1 %>% add_surface(showscale=FALSE)

fig2 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun1, scene='scene2') 
fig2 <- fig2 %>% add_surface(showscale=FALSE)

fig3 <- plot_ly(x = timefine, y = timefine,z = ~P_fun1, scene='scene3') 
fig3 <- fig3 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR1 <- subplot(fig1, fig2, fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range=c(-10,1200),
                                              title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                   title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                            title = "Phe"),
                                    aspectmode='cube'))

fig_RR1
```

2.  Fit the smoothed data on the original time points.

We interpolate the smoothed data at the original measurement points.

```{r}
trait_hat_list <- list()

for (i in 1:N){
  inter_trait <- convert_to_basisfunctions(t = timefine, eigenvecs = aligned_mass_curve[,i],
                                   tout = age_list_new[[i]])
  trait_hat_list[[i]] <- inter_trait
}

df_mass2 <- data.frame(id = id, trait = unsplit(trait_hat_list, id), phi_mass)

fmmForm2 <- trait ~ -1 + df_mass2$phi1 + df_mass2$phi2 + 
  (-1 + df_mass2$phi1 + df_mass2$phi2 | df_mass2$id) + 
  (-1 + df_mass2$phi1 + df_mass2$phi2 | df_mass2$id) 

system.time(
  ff2 <- fit_genetic_fmm(fmmForm2, df_mass2, A, 2)
) # user   system  elapsed

summary(ff2)
isSingular(ff2)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width = "150%"}
VC2 <- VarCorr(ff2)
CG2 <- VC2[["df_mass2.id"]]
CE2 <- VC2[["df_mass2.id.1"]]

### Convert to genetic covariance function
CG_fun2 <- pcs_mass %*% CG2 %*% t(pcs_mass)
### environmental covariance function
CE_fun2 <- pcs_mass %*% CE2 %*% t(pcs_mass)
### Phenotypic covariance function
P_fun2 <- CG_fun2 + CE_fun2

# individual plots
fig4 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun2, scene='scene1') 
fig4 <- fig4 %>% add_surface(showscale=FALSE)

fig5 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun2, scene='scene2') 
fig5 <- fig5 %>% add_surface(showscale=FALSE)

fig6 <- plot_ly(x = timefine, y = timefine,z = ~P_fun2, scene='scene3') 
fig6 <- fig6 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR2 <- subplot(fig4, fig5, fig6) 
fig_RR2 <- fig_RR2 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time"),
                                   zaxis=list(title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(range=c(-10,1200),
                                            title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"), 
                                    zaxis=list(range=c(-10,1200),
                                            title = "Phe"),
                                    aspectmode='cube'))

fig_RR2
```

3. Fit the smoothed data on the dense time grid.

```{r}
### Reform the aligned data to a dataframe for model fitting
subjectID <- rep(unique(id), each=100)
trait_pred <- c(aligned_mass_curve)
basis1 <- rep(pcs_mass[,1], times = 873)
basis2 <- rep(pcs_mass[,2], times = 873)
new_df <- data.frame(subjectID, trait_pred, basis1, basis2)
names(new_df) <- c("subjectID", "trait_hat", "basis1", "basis2")

fmeFormula <- trait_hat ~ -1 + new_df$basis1 + new_df$basis2 + 
  (-1 + new_df$basis1 + new_df$basis2 | new_df$subjectID) + 
  (-1 + new_df$basis1 + new_df$basis2 | new_df$subjectID)

system.time(
  fit_sameBasis <- fit_genetic_fmm(formula= fmeFormula, data=new_df, A = A, nbasis = 2)
) 
 # user   system   elapsed

summary(fit_sameBasis)
isSingular(fit_sameBasis)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width = "150%"}
VC3 <- VarCorr(fit_sameBasis)
CG3 <- VC3[["new_df.subjectID"]]
CE3 <- VC3[["new_df.subjectID.1"]]

### Convert to genetic covariance function
CG_fun3 <- pcs_mass %*% CG3 %*% t(pcs_mass)
### environmental covariance function
CE_fun3 <- pcs_mass %*% CE3 %*% t(pcs_mass)
### Phenotypic covariance function
P_fun3 <- CG_fun3 + CE_fun3

# individual plots
fig7 <- plot_ly(x = timefine, y = timefine, z = ~CG_fun3, scene='scene1') 
fig7 <- fig7 %>% add_surface(showscale=FALSE)

fig8 <- plot_ly(x = timefine, y = timefine,z = ~CE_fun3, scene='scene2') 
fig8 <- fig8 %>% add_surface(showscale=FALSE)

fig9 <- plot_ly(x = timefine, y = timefine,z = ~P_fun3, scene='scene3') 
fig9 <- fig9 %>% add_surface(showscale=FALSE)


# subplot and define scene
fig_RR3 <- subplot(fig7, fig8, fig9) 
fig_RR3 <- fig_RR3 %>% layout(title = "Covariance Plot",
                      scene = list(domain=list(x=c(0,0.32),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range=c(-10,1200),
                                              title = "Gen"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.34,0.66),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                             title = "Env"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.68,1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range=c(-10,1200),
                                             title = "Phe"),
                                    aspectmode='cube'))

fig_RR3
```
