---
title: "Smoothing Revisted and Model Fitting"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import library

```{r warning = FALSE, message = FALSE}
library(Matrix)
library(fdasrvf)
library(lme4)
library(pedigreemm)
library(ggplot2)
library(plotly)
```

## Import data and rescale time to unit interval.

```{r}
TRFUN25PUP4 = read.delim("TRFUN25PUP4.DAT",header = FALSE)
names(TRFUN25PUP4)<-c("id","sire","dam","trait","x")
df <- data.frame(TRFUN25PUP4)

FirstUniqueIdPos <- which(duplicated(df$id) == FALSE)
N = length(FirstUniqueIdPos) # N = 873 subjects
n = length(df$id) # n = 6860 observations
age_list <- split(df$x,df$id)
trait_list <- split(df$trait,df$id)

age_list_new <- list()
for (i in 1:N){
  age_list_new[[i]] = (age_list[[i]]-min(age_list[[i]]))/(max(age_list[[i]])-min(age_list[[i]]))
}

df$x_rescaled <- unsplit(age_list_new,df$id)
```

## Plot the raw data

```{r}
par(mfrow=c(1,2))
plot(c(0,25), c(0,400), xlab="Time", ylab="Mass", main="Growth Plot", type="n")
for (i in 1:N){
  lines(age_list[[i]], trait_list[[i]],type="l", col=i)
}

plot(c(0,25), c(0,3), type = 'n',xlab = 'Time', 
     ylab = 'Mass', main = 'Log Growth Plot ')
for (i in 1:N) {
  lines(age_list[[i]], log10(trait_list[[i]]), col = i)
}
```

# Data smoothing

Use penalised smoothing spline as basis functions. Here we take two ways to smooth the data. First, smooth body mass on its original. Second, smooth growth curves on the logarithmic scale and then take exponential to recover body mass. This imposes positive smoothing (same methodology as `smooth.pos()` in the **fda** package). We also report the smoothing parameter $\lambda$ selected by GCV.

```{r}
agefine <- seq(0,1,length=100) # create a fine time grid
mass_hat <- matrix(0,100,N) # store smooth growth curves on original scale
logmass_hat <- matrix(0,100,N) # store smooth log growth curves
pred_mass <- matrix(0,100,N) # store the smoothed mass recovered by taking exponential
lambda_mass <- rep(0,N) # smoothing parameter used for each growth curve
lambda_logmass <- rep(0,N) # smoothing parameter used for each log growth curve
for (i in 1:N){
  ss_mass <- smooth.spline(age_list_new[[i]], trait_list[[i]], cv=FALSE, 
                           all.knots=TRUE)
  ss_logmass <- smooth.spline(age_list_new[[i]], log10(trait_list[[i]]), cv=FALSE,
                              all.knots=TRUE) # all distinct points as knots
  mass_hat[,i] <- predict(ss_mass,agefine)$y
  logmass_hat[,i] <- predict(ss_logmass, agefine)$y
  pred_mass[,i] <- 10^(predict(ss_logmass,agefine)$y)
  lambda_mass[i] <- ss_mass$lambda
  lambda_logmass[i] <- ss_logmass$lambda
}
```

Some smoothed curves have negative values near $t=0$, let us plot them and the corresponding original data points.

```{r}
neg_cols <- unique(which(mass_hat < 0, arr.ind = TRUE)[,2])
length(neg_cols) 
neg_mass<- mass_hat[,neg_cols] # matrix with each column represents wrongly smoothed growth curves
colnames(neg_mass) <- as.character(neg_cols)
```

There are 77 questionable smoothed growth curves. On each plot, black points represent the data measurements; blue curve represents smoothing on the original scale; red curve represents smoothing by imposing positive constraint.

```{r include=FALSE}
par(mfrow=c(3,3))
for (i in 1:length(neg_cols)){
  plot(age_list_new[[neg_cols[i]]], trait_list[[neg_cols[i]]], type="p", xlab="Time", ylab="Mass", main=paste("Subject",neg_cols[i]))
  lines(agefine,mass_hat[,neg_cols[i]],type="l", col="blue")
  lines(agefine, pred_mass[,neg_cols[i]],type="l", col="red")
}
```

```{r echo=FALSE}
par(mfrow=c(2,2))
demo <- c(16, 46, 60, 534)
for (i in 1:length(demo)){
  plot(age_list_new[[demo[i]]], trait_list[[demo[i]]], type="p", xlab="Time", ylab="Mass", main=paste("Subject",demo[i]))
  lines(agefine,mass_hat[,demo[i]],type="l", col="blue")
  lines(agefine, pred_mass[,demo[i]],type="l", col="red")
}
```

```{r}
## Check the largest and smallest number of measurements per subject
##Initialize variables to hold the maximum and minimum lengths
max_length <- -Inf
min_length <- Inf

## Iterate through each list in trait_list
for (i in 1:length(trait_list)) {
  current_length <- length(trait_list[[i]])
  if (current_length > max_length) {
    max_length <- current_length
  }
  if (current_length < min_length) {
    min_length <- current_length
  }
}

print(paste("Largest number of elements in a list:", max_length))
print(paste("Smallest number of elements in a list:", min_length))
```

Sampling points are not taken at fixed times as they vary in number and location. The largest number of measurements per individual is 14 and the smallest is 5. Mass is measured more frequently in the last days of the time period. The irregular sampling points lead to two problems: for subjects with fewer measurements, GCV will select very large $\lambda$, and fitted curves approach to standard linear regression to the data (e.g. Subject 16, Subject 60); for subjects with more sparse measurements around the starting period, the fitted curves have negative values near $t=0$ (e.g. Subject 5, Subject 46).

If we compare the two methods to smooth the growth curves, we find that positive smoothing performs better in general. So we continue with positive smoothing and manually adjust the smoothing parameter $\lambda$ for subjects where the initial smoothing results are unsatisfactory, i.e. Subject 534 with $\lambda_{534} = 93480.4$.

Most smoothing parameter $\lambda$s selected by GCV are quite small. Let us restrict $\lambda \le 10^{-4}$, and re-smooth growth curves.

```{r include=FALSE}
large_lam <- which(lambda_logmass > 1e-4)
for (i in 1:length(large_lam)){
  plot(age_list_new[[large_lam[i]]], trait_list[[large_lam[i]]], type="p", xlab="Time", ylab="Mass", main=paste("Subject",large_lam[i]))
  lines(agefine, pred_mass[,large_lam[i]],type="l", col="red")
}

## change lambda to 1e-4
mass_adjusted <- matrix(0,100, length(large_lam))
for ( i in 1: length(large_lam)){
  ss_new <- smooth.spline(age_list_new[[large_lam[i]]], log(trait_list[[large_lam[i]]]), all.knots = TRUE, lambda = 1e-4)
  mass_adjusted[,i] <- exp(predict(ss_new, agefine)$y)
}
colnames(mass_adjusted) <- as.character(large_lam)
```

```{r include=FALSE}
par(mfrow=c(3,3))
for (i in 1:length(large_lam)){
  plot(age_list_new[[large_lam[i]]], trait_list[[large_lam[i]]], type="p", xlab="Time", ylab="Mass", main=paste("Subject",large_lam[i]))
  lines(agefine, mass_adjusted[,i],type="l", col="red")
}
```


```{r}
mass_smoothed <-list()
pred_mass_fine <- matrix(0,100,N) # store the smoothed mass predicted on the dense grid
pred_logmass_fine <- matrix(0,100,N) # store the smoothed logmass
lam <- rep(0,N) # smoothing parameter used for each log growth curve

for (i in 1:N) {
  ss_logmass <- smooth.spline(age_list_new[[i]], log10(trait_list[[i]]), cv=FALSE, all.knots=TRUE)
  
  # Check if lambda is greater than 1e-4
  if (ss_logmass$lambda > 1e-4) {
    # Redo smoothing with lambda set to 1e-4
    ss_logmass <- smooth.spline(age_list_new[[i]], log10(trait_list[[i]]), lambda=1e-4, all.knots=TRUE)
  }
  
  mass_smoothed[[i]] <- 10^(ss_logmass$y)
  pred_logmass_fine[,i] <- predict(ss_logmass, agefine)$y
  pred_mass_fine[,i] <- 10^(predict(ss_logmass, agefine)$y) ## predict smoothed data on a regular dense grid
  lam[i] <- ss_logmass$lambda
}
```

```{r}
par(mfrow=c(1,2))

plot(c(0,1), c(0,400), xlab="Time", ylab="Mass", main="Growth Plot (rescaled time)", type="n")
for (i in 1:N){
  lines(age_list_new[[i]], trait_list[[i]],type="l", col=i)
}

matplot(agefine, pred_mass_fine, col=1:N, type = "l", xlab="Time", ylab="Mass", main="Smoothed Growth Plot")
```

## Curve registration

```{r message=FALSE, warning=FALSE}
aligned_mass_process <- time_warping(f=pred_mass_fine, time=agefine)
aligned_mass_curve <- aligned_mass_process$fn
aligned_mean <- aligned_mass_process$fmean
warping_funs <- aligned_mass_process$warping_functions


par(mfrow=c(1,3))
plot(c(0,1), c(0,1), type = 'n',xlab = 'Time', 
     ylab = 'warping functions')
for (i in 1:N){
  lines(agefine, warping_funs[,i], type="l" , col=i)
}


plot(c(0,1), c(0,400), type = 'n',xlab = 'Time', 
     ylab = 'Mass', main = 'Aligned Growth Plot')
for (i in 1:N){
  lines(agefine, aligned_mass_curve[,i], type="l",col=i)
}
lines(agefine, aligned_mean, lwd = 3.0, col="red")
legend("bottomright", legend="mean", lwd=3.0, col="red")

matplot(agefine, pred_mass_fine, col=1:N, type = "l", xlab="Time", ylab="Mass", main="Smoothed Growth Plot")
```

## Functional Principal Component Analysis

```{r}
fpcaobj_mass <- prcomp(x=t(aligned_mass_curve), retx = TRUE, center = TRUE, rank. = 3)
eigen_mass <- fpcaobj_mass$rotation # eigen vectors

par(mfrow=c(3,1))
for (i in 1:3) {
  plot(agefine, eigen_mass[, i], type = "l", 
       xlab = "Time", ylab = "",
       main = paste("Growth Curve Principal Component", i))
}

## Test for orthogonality
eigen_mass[,1] %*% eigen_mass[,2] 
eigen_mass[,1] %*% eigen_mass[,3] 
eigen_mass[,2] %*% eigen_mass[,3] 
```
