---
title: "Simulation(PCs Example)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(Matrix)
library(MASS)
library(fda)
library(lme4)
library(fdasrvf)
library(pedigreemm)
library(mvnfast)
library(ggplot2)
library(plotly)
library(fBasics)
library(roahd)
library(robustlmm)
library(lme4pureR)
library(latex2exp)

convert_to_basisfunctions <- function(t, eigenvecs, tout) 
{
  #'This function converts an eigenvector to an eigenfunction using linear interpolation.
  #'
  #'@param t a numeric vector containing the fine time grid points used to compute the principal components.
  #'@param eigenvecs vector or matrix of eigenvectors obtained from FPCA.
  #'@param tout an optional set of numeric values specifying where interpolation is to take place.
  #'@return matrix where each column represents an eigenfunction of time.
  
  # Check if eigenvecs is a vector or matrix
  if (is.vector(eigenvecs)) {
    eigen_functions <- approx(x = t, y = eigenvecs, xout = tout)$y
  } else {
    eigen_functions <- matrix(0, nrow = length(tout), ncol = ncol(eigenvecs))
    for (i in 1:ncol(eigenvecs)) {
      eigen_functions[,i] <- approx(x = t, y = eigenvecs[,i], xout = tout)$y
    }
  }
  
  return(eigen_functions)
}

fit_genetic_fmm <- function(formula, data, A, nbasis, control = lmerControl()) {
  #' This function uses the lme4 package to fit a linear mixed-effect model to genetic data, 
  #' with a specified additive genetic relationship matrix A.
  #' In this particular format, we fit both fixed effects and random effects using the same
  #' basis functions (principal components obtained from running FPCA).
  #'
  #' @param formula a two-sided linear formula object describing both the fixed-effects
  #' and random-effects of the model (as the same form used in lmer).
  #' @param data a data frame containing the variables named in formula.
  #' @param A a sparse matrix: an additive genetic relationship matrix 
  #' which models the genetic relationship in the dataset.
  #' @param nbasis numeber: number of basis used to fit the model
  #' @param control lmerControl object: control parameters for the optimizer.
  #' @return returns a fitted mixed-effect model
  
  # Load required packages
  library(lme4)
  library(Matrix)
  
  # Cholesky decomposition of A
  LA <- as(t(chol(A)), "sparseMatrix")
  I_p <- as(diag(nbasis), "sparseMatrix")
  MA <- kronecker(LA, I_p) # used to update the genetic design matrix Z_G = ZM
  
  # Define the mixed-model formula
  fmmParsedForm <- lFormula(formula=formula, data = data, control = control)
  
  # Compute the random-effect matrix
  Z_pre <- t(fmmParsedForm$reTrms$Zt)
  ZE <- Z_pre[, 1:dim(MA)[1]] # environmental random-effect matrix
  ZG <- Z_pre[, 1:dim(MA)[1]] %*% MA # update the genetic-random effect matrix
  Z <- cbind(ZG, ZE) # the updated random effect design matrix
  
  # Modularisation
  fmmParsedForm$reTrms$Zt <- t(Z) # Update Z in the reTrms term
  fmmDevFun <- do.call(mkLmerDevfun, fmmParsedForm) # update the objective function
  
  fmmOptimize <- optimizeLmer(devfun = fmmDevFun, control = control) # update the optimisation module
  
  # Return the mixed-effect model
  fmm <- mkMerMod(rho = environment(fmmDevFun), opt = fmmOptimize, reTrms = fmmParsedForm$reTrms, fr = fmmParsedForm$fr)
  
  return(fmm)
}
```


# Simulation results (using PCs)

Generate data: using 5 spline basis; 873 individual curves, 7 measurements per individual, residual variance = 15.

Fit smoothed data using 6 PCs for estimating the fixed-effect and 5 PCs for estimating the random effects.

```{r}
## set up true covariance
N <- 873 # total number of individuals
n <- 7 # number of measurements per individual
nbasis <- 5 # number of basis
ngroups <- 50

### use b-spline basis
basisObj <- create.bspline.basis(c(0,1), nbasis = nbasis, norder = 4)

### genetic covariance matrix
C_gen <- matrix(c(750, 10 ,130, 80, 250,
                  10, 800, 30, 15, 40,
                  130, 30, 700, 50, 130,
                  80, 15, 50, 420, 50,
                  250, 40, 130, 50, 330), nrow = 5, byrow = T)
### environmental covariance matrix
C_env <- matrix(c(750, 10 ,130, 80, 250,
                  10, 800, 30, 15, 40,
                  130, 30, 700, 50, 130,
                  80, 15, 50, 420, 50,
                  250, 40, 130, 50, 330), nrow = 5, byrow = T)
### residual variance
sigma2 <- 15

### model reform
time_rang <- seq(0,1,length=n) # time points t_j
timefine <- seq(0,1, length =100)
basis <- eval.basis(time_rang, basisObj)

CG_fun_true <- basis %*% C_gen %*% t(basis) # true genetic covariance function
CE_fun_true <- basis %*% C_env %*% t(basis) # true environmental covariance function
```

```{r include=FALSE}
files <- list.files(pattern = "\\.Rdata$")

for (file in files) {
  load(file)
}

load("TC_fixed_effect.Rdata")
```

## Simulation result

Error on fixed-effects:

```{r}
par(mfrow = c(1,2))
FefErr_list <- list(ferr_5715)

# Generate the boxplot
boxplot(FefErr_list,
        xlab = "",
        ylab = "Error",
        main = "Fixed-Effect Error")

```

```{r}
load("Y_5715.Rdata")
uniqueIds <- seq(1,N, length=N)
id <- rep(uniqueIds, each=n)

Y20_list <- split(Y_5715[,20], id)
Y31_list <- split(Y_5715[,31], id)

par(mfrow = c(1,1))
plot(c(0,1), c(-150, 350), type = "n", xlab = "t", ylab = "", main = "Group ID: 20")
for (i in 1:N){
  lines(time_rang, Y20_list[[i]], col = i, type = "l")
}

par(mfrow = c(1,1))
plot(c(0,1), c(-150, 350), type = "n", xlab = "t", ylab = "", main = "Group ID: 31")
for (i in 1:N){
  lines(time_rang, Y31_list[[i]], col = i, type = "l")
}

par(mfrow = c(1,1))
plot(c(0,1), c(0, 290), type = "n", xlab = "t", ylab = "", main = "Fixed-Effect")
lines(time_rang, f_true, col = "red", lty = "solid")
lines(time_rang, fef_5pc715[,20], col = "black", lty = "dashed")
lines(time_rang, fef_5pc715[,31], col = "blue", lty = "dashed")
legend("bottomright", legend = c("true", "estimated ID 20", "estimated ID 31"), lty = c("solid","dashed", "dashed"), col = c("red","black", "blue"), bty = "n", cex = 0.8)
```

Error on covariance functions:

GenCov outlier ID: 14 15 23 43

EnvCov outlier ID: 15 23 31 43

```{r}
sfitID <- which(sfit_5pc715)
sfitID
```


```{r include=FALSE}
Y15_list <- split(Y_5715[,15], id)
Y23_list <- split(Y_5715[,23], id)
Y43_list <- split(Y_5715[,43],id)
Y11_list <- split(Y_5715[,11], id)
Y25_list <- split(Y_5715[,25], id)

Y41_list <- split(Y_5715[,41],id)
Y48_list <- split(Y_5715[,48],id)
```

## Group ID 41 (smallest GenErr) 

```{r echo=FALSE}
par(mfrow = c(1,1))
plot(c(0,1), c(-150, 350), type = "n", xlab = "t", ylab = "", main = "Group ID: 41")
for (i in 1:N){
  lines(time_rang, Y41_list[[i]], col = i, type = "l")
}
```

```{r}
f_true <- convert_to_basisfunctions(timefine, fixed_effect, time_rang)
par(mfrow = c(1,1))
plot(c(0,1), c(0, 290), type = "n", xlab = "t", ylab = "", main = TeX("Fixed-Effect ($\\sigma^2=15$)"))
lines(time_rang, f_true, col = "red", lty = "solid")
lines(time_rang, fef_5pc715[,41], col = "black", lty = "dashed")
legend("bottomright", legend = c("true", "estimated ID 41"), lty = c("solid","dashed"), col = c("red","black"), bty = "n", cex = 0.8)
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,41], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,41], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 41",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r}
pcs_5pc715[,,41]
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,48], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,48], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 48",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r}
pcs_5pc715[,,48]
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,11], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,11], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 11 (singular)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,23], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,23], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 23 (singular)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,25], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,25], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 25 (singular)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,43], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,43], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 43 (singular)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r}
pcs_5pc715[,,43]
```


```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,15], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,15], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 15 (outlier)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,14], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,14], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 14 (Outlier)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,25], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,25], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 25 (Outlier)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```

```{r message=FALSE, echo=FALSE, warning=FALSE, out.width="150%"}
fig1 <- plot_ly(showscale=FALSE, scene='scene1') 
fig1 <- fig1 %>% add_surface(z = ~CG_fun_true, x = time_rang, y = time_rang)

fig2 <- plot_ly(showscale = FALSE, scene='scene2') 
fig2 <- fig2 %>% add_surface(z = ~CG_fun_5pc715[,,31], x = time_rang, y = time_rang)

fig3 <- plot_ly(showscale = FALSE, scene='scene3') 
fig3 <- fig3 %>% add_surface(z = ~CE_fun_5pc715[,,31], x = time_rang, y = time_rang)

fig_RR1 <- subplot(fig1, fig2,fig3) 
fig_RR1 <- fig_RR1 %>% layout(title = "Group ID 31 (Outlier)",
                      scene = list(domain=list(x=c(0,0.30),y=c(0.25,1)),
                                   xaxis=list(title = "time"),
                                   yaxis =list(title = "time") , 
                                   zaxis=list(range = c(-100,900),title = "True Cov"),
                                   aspectmode='cube'),
                      scene2 = list(domain=list(x=c(0.35,0.65),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Gen"),
                                    aspectmode='cube'),
                      scene3 = list(domain=list(x=c(0.70,0.1),y=c(0.25,1)),
                                    xaxis=list(title = "time"),
                                    yaxis =list(title = "time"),
                                    zaxis=list(range = c(-100,900),title = "Estimated Env"),
                                    aspectmode='cube')
                      )

fig_RR1
```
